---
title: "Bulk RNA-seq preprocessing and quality control"
author: "Andreu Odena"
date: "2026-08-24"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "svg",
  fig.align = "center"
)
```

# Objective

This analysis prepares the baseline bulk RNA-seq expression matrix used in the study.
It performs sample-level quality control, removes genes with low counts, applies TMM normalization and generates a log2-CPM matrix for downstream analyses.

It generates supplementary quality-control figures:

1. Library size for each PDX model
2. Clinicopathological-subtype composition within each sequencing batch
3. PCA of the normalized expression profiles.

# Packages and input files

The analysis should be run from the root of the GitHub repository. Raw data are
expected in the local `data` directory and are not included in the public
repository.

```{r packages-and-paths}
library(edgeR)
library(ggplot2)

counts_path <- file.path("data", "260824_Counts.csv")
metadata_path <- file.path("data", "260824_Metadata.csv")
figure_dir <- file.path("figures", "supplementary_RNAseq_QC")

dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
```

# Load the data

```{r load-data}
counts_input <- read.csv(
  counts_path,
  sep = ";",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

metadata <- read.csv(
  metadata_path,
  sep = ";",
  check.names = FALSE,
  stringsAsFactors = FALSE,
  fileEncoding = "UTF-8-BOM"
)

gene_names <- trimws(as.character(counts_input$gene_name))
counts_matrix <- as.matrix(counts_input[, -1, drop = FALSE])
storage.mode(counts_matrix) <- "numeric"
rownames(counts_matrix) <- gene_names

metadata$samples <- trimws(as.character(metadata$samples))

# Reorder the metadata to match the count-matrix columns exactly.
metadata <- metadata[
  match(colnames(counts_matrix), metadata$samples),
  ,
  drop = FALSE
]

metadata$Batch <- factor(metadata$Batch)
metadata$Subtype_2 <- factor(
  metadata$Subtype_2,
  levels = c("ER+/HER2-", "ER+/HER2+", "TNBC")
)

cat("Raw count-matrix dimensions:", dim(counts_matrix), "\n")
cat("Number of PDX models:", ncol(counts_matrix), "\n")
```

# Sample-level exploration

## Library size

Library size was calculated as the total number of mapped reads assigned to genes in each sample before expression filtering.

```{r library-size, fig.width=7, fig.height=11}
library_size_df <- data.frame(
  sample = colnames(counts_matrix),
  library_size_million = colSums(counts_matrix) / 1e6,
  Batch = metadata$Batch,
  Subtype = metadata$Subtype_2,
  stringsAsFactors = FALSE
)

library_size_df <- library_size_df[
  order(library_size_df$library_size_million),
  ,
  drop = FALSE
]

library_size_df$sample <- factor(
  library_size_df$sample,
  levels = library_size_df$sample
)

batch_colours <- c(
  "SERRAVIO_03" = "#4C72B0",
  "SERRAVIO_04" = "#DD8452",
  "SERRAVIO_09" = "#55A868"
)

p_library_size <- ggplot(
  library_size_df,
  aes(x = sample, y = library_size_million, fill = Batch)
) +
  geom_col(width = 0.8) +
  coord_flip() +
  scale_fill_manual(values = batch_colours) +
  labs(
    x = NULL,
    y = "Library size (million reads)",
    fill = "Sequencing batch"
  ) +
  theme_classic(base_size = 11) +
  theme(axis.text.y = element_text(size = 5))

p_library_size

ggsave(
  file.path(figure_dir, "Figure_library_size.svg"),
  p_library_size,
  width = 7,
  height = 11
)
```

## Subtype composition by sequencing batch

This plot checks whether the proportions of the main clinicopathological subtypes differ across RNA-sequencing batches.

```{r batch-subtype-composition, fig.width=7, fig.height=5}
batch_subtype_df <- as.data.frame(
  table(
    Batch = metadata$Batch,
    Subtype = metadata$Subtype_2
  )
)

colnames(batch_subtype_df)[3] <- "n_models"

subtype_colours <- c(
  "ER+/HER2-" = "#4C72B0",
  "ER+/HER2+" = "#C65CCB",
  "TNBC" = "#F8766D"
)

p_batch_subtype <- ggplot(
  batch_subtype_df,
  aes(x = Batch, y = n_models, fill = Subtype)
) +
  geom_col(
    position = "fill",
    colour = "white",
    linewidth = 0.3
  ) +
  scale_y_continuous(
    labels = function(x) paste0(round(100 * x), "%")
  ) +
  scale_fill_manual(values = subtype_colours, drop = FALSE) +
  labs(
    x = "Sequencing batch",
    y = "Proportion of PDX models",
    fill = "Clinicopathological subtype"
  ) +
  theme_classic(base_size = 11)

p_batch_subtype

ggsave(
  file.path(figure_dir, "Figure_batch_subtype_composition.svg"),
  p_batch_subtype,
  width = 7,
  height = 5
)
```

# Low-expression filtering

Following the procedure used in class, genes were retained when they had more
than 10 reads in at least 15 PDX models. The threshold of 15 models was fixed
before downstream analysis and corresponds to approximately 19% of the 79-model cohort.

```{r expression-filtering}
minimum_count <- 10
minimum_samples <- 15

keep_gene <- rowSums(counts_matrix > minimum_count) >= minimum_samples
counts_filtered <- counts_matrix[keep_gene, , drop = FALSE]

cat("Genes before filtering:", nrow(counts_matrix), "\n")
cat("Genes after filtering:", nrow(counts_filtered), "\n")
```

# TMM normalisation

TMM normalisation was performed with `edgeR` to account for differences in
library composition. Normalised log2 counts per million were then calculated
for visualisation and downstream sample-level analyses.

```{r tmm-normalisation}
dge <- DGEList(counts = counts_filtered)
dge <- calcNormFactors(dge, method = "TMM")

expression_log2cpm <- cpm(
  dge,
  log = TRUE,
  prior.count = 1
)

```

# PCA

PCA was performed on the transposed TMM-normalised log2-CPM matrix. Genes were
centred and scaled. The analysis is unsupervised and was not batch-corrected; batch and subtype were used only as plot annotations.

```{r pca, fig.width=8, fig.height=6}
pca_result <- prcomp(
  t(expression_log2cpm),
  center = TRUE,
  scale. = TRUE
)

variance_explained <- 100 * (
  pca_result$sdev^2 / sum(pca_result$sdev^2)
)

pca_df <- data.frame(
  sample = rownames(pca_result$x),
  PC1 = pca_result$x[, "PC1"],
  PC2 = pca_result$x[, "PC2"],
  Batch = metadata$Batch,
  Subtype = metadata$Subtype_2,
  stringsAsFactors = FALSE
)

p_pca <- ggplot(
  pca_df,
  aes(
    x = PC1,
    y = PC2,
    colour = Subtype,
    shape = Batch
  )
) +
  geom_point(size = 3.2, alpha = 0.9) +
  scale_colour_manual(values = subtype_colours, drop = FALSE) +
  scale_shape_manual(
    values = c(
      "SERRAVIO_03" = 16,
      "SERRAVIO_04" = 17,
      "SERRAVIO_09" = 15
    )
  ) +
  labs(
    x = paste0("PC1 (", round(variance_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(variance_explained[2], 1), "%)"),
    colour = "Clinicopathological subtype",
    shape = "Sequencing batch"
  ) +
  theme_classic(base_size = 11)

p_pca

ggsave(
  file.path(figure_dir, "Figure_PCA_subtype_batch.svg"),
  p_pca,
  width = 8,
  height = 6
)
```

# Reproducibility information

```{r session-information}
sessionInfo()
```
