---
title: "PAM50 intrinsic-subtype classification"
author: "Andreu Odena"
date: "2026-09-08"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "svg",
  fig.align = "center"
)
```

# Objective

This analysis assigns a PAM50 intrinsic subtype to each baseline PDX model from
TMM-normalised bulk RNA-seq expression. The Parker PAM50 classification is used
as the principal subtype call, while the `genefu.scale` result is retained as a
complementary classification. The analysis also generates a supplementary PCA
coloured by the calculated PAM50 subtype and shaped by sequencing batch.

# Packages and paths

```{r packages-and-paths}
library(edgeR)
library(BreastSubtypeR)
library(SummarizedExperiment)
library(S4Vectors)
library(ggplot2)

counts_path <- file.path("data", "260824_Counts.csv")
metadata_path <- file.path("data", "260824_Metadata.csv")

results_dir <- file.path("results", "PAM50")
figure_dir <- file.path("figures", "supplementary_PAM50")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
```

# Load and preprocess the RNA-seq data

Genes were retained when they had more than 10 reads in at least 15 PDX models.
TMM normalisation and log2-CPM transformation were performed with `edgeR`, using
the same preprocessing settings as the main bulk RNA-seq analysis.

```{r load-and-normalise-data}
counts_input <- read.csv(
  counts_path,
  sep = ";",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

metadata <- read.csv(
  metadata_path,
  sep = ";",
  check.names = FALSE,
  stringsAsFactors = FALSE,
  fileEncoding = "UTF-8-BOM"
)

gene_names <- trimws(as.character(counts_input$gene_name))
counts_matrix <- as.matrix(counts_input[, -1, drop = FALSE])
storage.mode(counts_matrix) <- "numeric"
rownames(counts_matrix) <- gene_names

metadata$samples <- trimws(as.character(metadata$samples))
metadata <- metadata[
  match(colnames(counts_matrix), metadata$samples),
  ,
  drop = FALSE
]

keep_gene <- rowSums(counts_matrix > 10) >= 15
counts_filtered <- counts_matrix[keep_gene, , drop = FALSE]

dge <- DGEList(counts = counts_filtered)
dge <- calcNormFactors(dge, method = "TMM")

expression_log2cpm <- cpm(
  dge,
  log = TRUE,
  prior.count = 1
)
```

# Map gene symbols to Entrez identifiers

`BreastSubtypeR` requires expression features annotated with gene symbol and
Entrez identifier. When several expression features map to the same PAM50
feature, the feature with the highest expression is retained by the package
mapping procedure.

```{r gene-annotation}
data("Gene.ID.ann", package = "BreastSubtypeR")

gene_symbols <- trimws(rownames(expression_log2cpm))
annotation_symbols <- trimws(as.character(Gene.ID.ann$HGNC))
annotation_entrez <- trimws(as.character(Gene.ID.ann$EntrezGene))

gene_match <- match(
  toupper(gene_symbols),
  toupper(annotation_symbols)
)

entrez_ids <- annotation_entrez[gene_match]
valid_genes <- !is.na(entrez_ids) & entrez_ids != ""

expression_for_pam50 <- expression_log2cpm[
  valid_genes,
  ,
  drop = FALSE
]

gene_symbols <- gene_symbols[valid_genes]
entrez_ids <- entrez_ids[valid_genes]

feature_row_names <- make.unique(gene_symbols)
rownames(expression_for_pam50) <- feature_row_names

row_annotation <- S4Vectors::DataFrame(
  probe = gene_symbols,
  SYMBOL = gene_symbols,
  ENTREZID = entrez_ids,
  row.names = feature_row_names
)
```

# Create the annotated expression object

Each PDX model is treated as an independent sample. Clinicopathological subtype
and sequencing batch are retained as sample annotations but are not used to
assign the PAM50 subtype.

```{r annotated-expression-object}
col_annotation <- S4Vectors::DataFrame(
  PatientID = metadata$samples,
  samples = metadata$samples,
  Batch = as.character(metadata$Batch),
  Subtype_2 = as.character(metadata$Subtype_2),
  ER = ifelse(
    metadata$Subtype_2 %in% c("ER+/HER2-", "ER+/HER2+"),
    "ER+",
    "ER-"
  ),
  HER2 = ifelse(
    metadata$Subtype_2 == "ER+/HER2+",
    "HER2+",
    "HER2-"
  ),
  TN = ifelse(
    metadata$Subtype_2 == "TNBC",
    "TN",
    "nonTN"
  ),
  row.names = metadata$samples
)

se_object <- SummarizedExperiment::SummarizedExperiment(
  assays = list(
    counts = as.matrix(expression_for_pam50)
  ),
  rowData = row_annotation,
  colData = col_annotation
)
```

# PAM50 classification

The TMM-normalised log2-CPM matrix is mapped to the PAM50 features and classified
using the Parker original and `genefu.scale` methods implemented in
`BreastSubtypeR`.

```{r pam50-classification}
pam50_input <- BreastSubtypeR::Mapping(
  se_obj = se_object,
  RawCounts = FALSE,
  method = "max",
  impute = TRUE,
  verbose = TRUE
)

pam50_result <- BreastSubtypeR::BS_Multi(
  data_input = pam50_input,
  methods = c("parker.original", "genefu.scale"),
  Subtype = FALSE,
  hasClinical = FALSE
)

pam50_calls <- as.data.frame(pam50_result$res_subtypes)
pam50_calls$samples <- rownames(pam50_calls)

colnames(pam50_calls)[
  colnames(pam50_calls) == "parker.original"
] <- "PAM50_Parker"

colnames(pam50_calls)[
  colnames(pam50_calls) == "genefu.scale"
] <- "PAM50_Genefu_scale"

pam50_label_map <- c(
  "LumA" = "Luminal A",
  "LumB" = "Luminal B",
  "Her2" = "HER2-enriched",
  "Basal" = "Basal-like",
  "Normal" = "Normal-like"
)

pam50_calls$PAM50_label <- unname(
  pam50_label_map[
    as.character(pam50_calls$PAM50_Parker)
  ]
)

pam50_calls <- merge(
  metadata[, c("samples", "Batch", "Subtype_2")],
  pam50_calls,
  by = "samples",
  all.x = TRUE,
  sort = FALSE
)

pam50_calls <- pam50_calls[
  match(metadata$samples, pam50_calls$samples),
  ,
  drop = FALSE
]

write.csv(
  pam50_calls,
  file.path(results_dir, "PAM50_Parker_calls.csv"),
  row.names = FALSE,
  quote = TRUE
)
```

# PAM50-gene expression matrix

The TMM-normalised expression values for the PAM50 signature genes are exported
for the cohort heatmap and other downstream visualisations.

```{r pam50-expression-matrix}
data("BreastSubtypeRobj", package = "BreastSubtypeR")

pam50_signature <- BreastSubtypeRobj$genes.sig50

character_columns <- colnames(pam50_signature)[
  vapply(
    pam50_signature,
    function(current_column) {
      is.character(current_column) || is.factor(current_column)
    },
    logical(1)
  )
]

signature_overlap <- vapply(
  character_columns,
  function(current_column) {
    sum(
      toupper(as.character(
        pam50_signature[[current_column]]
      )) %in% toupper(gene_symbols)
    )
  },
  numeric(1)
)

pam50_gene_column <- character_columns[
  which.max(signature_overlap)
]

pam50_genes <- unique(
  toupper(as.character(
    pam50_signature[[pam50_gene_column]]
  ))
)

pam50_gene_indices <- which(
  toupper(gene_symbols) %in% pam50_genes
)

pam50_expression <- expression_for_pam50[
  pam50_gene_indices,
  ,
  drop = FALSE
]

write.csv(
  data.frame(
    gene_name = gene_symbols[pam50_gene_indices],
    pam50_expression,
    check.names = FALSE
  ),
  file.path(results_dir, "PAM50_expression_log2CPM.csv"),
  row.names = FALSE,
  quote = TRUE
)
```

# PCA annotated by PAM50 intrinsic subtype

PCA is performed using the 500 genes with the highest variance across the
TMM-normalised cohort. The calculated Parker PAM50 subtype is used only as a
colour annotation and sequencing batch is represented by point shape.

```{r pam50-pca, fig.width=8.5, fig.height=6.2}
gene_variance <- apply(
  expression_log2cpm,
  1,
  var,
  na.rm = TRUE
)

top_variable_genes <- names(
  sort(gene_variance, decreasing = TRUE)
)[seq_len(500)]

pca_expression <- expression_log2cpm[
  top_variable_genes,
  ,
  drop = FALSE
]

pca_result <- prcomp(
  t(pca_expression),
  center = TRUE,
  scale. = TRUE
)

explained_variance <- 100 * (
  pca_result$sdev^2 / sum(pca_result$sdev^2)
)

pca_data <- data.frame(
  samples = rownames(pca_result$x),
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  stringsAsFactors = FALSE
)

pam50_index <- match(
  pca_data$samples,
  pam50_calls$samples
)

metadata_index <- match(
  pca_data$samples,
  metadata$samples
)

pca_data$PAM50_label <- pam50_calls$PAM50_label[
  pam50_index
]
pca_data$Batch <- metadata$Batch[metadata_index]

pam50_order <- c(
  "Luminal A",
  "Luminal B",
  "HER2-enriched",
  "Basal-like",
  "Normal-like"
)

pca_data$PAM50_label <- factor(
  pca_data$PAM50_label,
  levels = pam50_order
)

pam50_colours <- c(
  "Luminal A" = "#4C78A8",
  "Luminal B" = "#2A6FBB",
  "HER2-enriched" = "#B455C5",
  "Basal-like" = "#D84A4A",
  "Normal-like" = "#7A8B8E"
)

batch_shapes <- c(
  "SERRAVIO_03" = 16,
  "SERRAVIO_04" = 17,
  "SERRAVIO_09" = 15
)

p_pca_pam50 <- ggplot(
  pca_data,
  aes(
    x = PC1,
    y = PC2,
    colour = PAM50_label,
    shape = Batch
  )
) +
  geom_point(size = 3.2, alpha = 0.9) +
  scale_colour_manual(
    values = pam50_colours,
    breaks = pam50_order,
    name = "PAM50 intrinsic subtype",
    drop = FALSE
  ) +
  scale_shape_manual(
    values = batch_shapes,
    name = "Sequencing batch"
  ) +
  labs(
    x = paste0(
      "PC1 (",
      round(explained_variance[1], 1),
      "%)"
    ),
    y = paste0(
      "PC2 (",
      round(explained_variance[2], 1),
      "%)"
    )
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

p_pca_pam50

ggsave(
  file.path(
    figure_dir,
    "PCA_TMM_log2CPM_by_PAM50.svg"
  ),
  p_pca_pam50,
  width = 8.5,
  height = 6.2
)
```

# Reproducibility information

```{r session-information}
sessionInfo()
```
