---
title: "Genomic-transcriptomic concordance analyses"
author: "Andreu Odena"
date: "2026-09-08"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Objective

This analysis first tests direct associations between selected copy-number
alterations and expression of the corresponding genes. It then calculates
model-level single-sample gene-set enrichment analysis (ssGSEA) scores for the
MSigDB Hallmark collection and tests associations between selected genomic
alterations and their corresponding transcriptional programs. Association
models are adjusted for the Parker PAM50 intrinsic subtype.

No figures are generated by this R Markdown.

# Packages and paths

The analysis should be run from the root of the GitHub repository. Raw count and
metadata files are expected in the local `data` directory. The PAM50 calls are
generated by `02_PAM50_classification.Rmd`.

```{r packages-and-paths}
library(edgeR)
library(limma)
library(GSVA)
library(msigdbr)
library(sandwich)
library(lmtest)

counts_path <- file.path("data", "260824_Counts.csv")
metadata_path <- file.path("data", "260824_Metadata.csv")
pam50_path <- file.path(
  "results",
  "PAM50",
  "PAM50_Parker_calls.csv"
)

results_dir <- file.path(
  "results",
  "genomic_transcriptomic_concordance"
)

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

# Load and normalise the RNA-seq data

Genes are retained when they have more than 10 counts in at least 15 PDX models.
TMM normalisation and log2-CPM transformation are performed using the same
settings as in the RNA-seq preprocessing analysis.

```{r load-and-normalise-data}
counts_input <- read.csv(
  counts_path,
  sep = ";",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

metadata <- read.csv(
  metadata_path,
  sep = ";",
  check.names = FALSE,
  stringsAsFactors = FALSE,
  fileEncoding = "UTF-8-BOM"
)

pam50_calls <- read.csv(
  pam50_path,
  check.names = FALSE,
  stringsAsFactors = FALSE,
  fileEncoding = "UTF-8-BOM"
)

gene_names <- trimws(as.character(counts_input$gene_name))
counts_matrix <- as.matrix(counts_input[, -1, drop = FALSE])
storage.mode(counts_matrix) <- "numeric"
rownames(counts_matrix) <- gene_names

metadata$samples <- trimws(as.character(metadata$samples))
pam50_calls$samples <- trimws(as.character(pam50_calls$samples))

metadata <- metadata[
  match(colnames(counts_matrix), metadata$samples),
  ,
  drop = FALSE
]

keep_gene <- rowSums(counts_matrix > 10) >= 15
counts_filtered <- counts_matrix[keep_gene, , drop = FALSE]

dge <- DGEList(counts = counts_filtered)
dge <- calcNormFactors(dge, method = "TMM")

expression_log2cpm <- cpm(
  dge,
  log = TRUE,
  prior.count = 1
)

rownames(expression_log2cpm) <- toupper(
  trimws(rownames(expression_log2cpm))
)

expression_log2cpm <- limma::avereps(
  expression_log2cpm,
  ID = rownames(expression_log2cpm)
)
```

# Calculate Hallmark ssGSEA scores

Human Hallmark gene sets are obtained from MSigDB through `msigdbr`. ssGSEA
scores are calculated from the TMM-normalised log2-CPM matrix using the `GSVA`
implementation with its standard weighting parameter and final score
normalisation.

```{r hallmark-gene-sets}
hallmark_genes <- msigdbr::msigdbr(
  species = "Homo sapiens",
  collection = "H"
)

hallmark_gene_sets <- split(
  toupper(hallmark_genes$gene_symbol),
  hallmark_genes$gs_name
)

hallmark_gene_sets <- lapply(
  hallmark_gene_sets,
  unique
)
```

```{r calculate-ssgsea-scores}
ssgsea_parameters <- GSVA::ssgseaParam(
  exprData = expression_log2cpm,
  geneSets = hallmark_gene_sets,
  minSize = 10,
  maxSize = 500,
  alpha = 0.25,
  normalize = TRUE
)

hallmark_ssgsea_scores <- GSVA::gsva(
  ssgsea_parameters,
  verbose = FALSE
)

ssgsea_output <- data.frame(
  samples = colnames(hallmark_ssgsea_scores),
  t(hallmark_ssgsea_scores),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

write.csv(
  ssgsea_output,
  file.path(results_dir, "PDX_Hallmark_ssGSEA_scores.csv"),
  row.names = FALSE,
  quote = TRUE
)
```

# Prepare genomic and PAM50 annotations

The Parker PAM50 calls calculated in the previous analysis are added to the
genomic metadata. Explicit `NO` or `WT` values define the reference group.
Other alterations are excluded when a specific genomic event is tested.

```{r prepare-annotations}
analysis_metadata <- merge(
  metadata,
  pam50_calls[, c("samples", "PAM50_label")],
  by = "samples",
  all.x = TRUE,
  sort = FALSE
)

analysis_metadata <- analysis_metadata[
  match(
    colnames(hallmark_ssgsea_scores),
    analysis_metadata$samples
  ),
  ,
  drop = FALSE
]

analysis_metadata$PAM50_label <- trimws(
  as.character(analysis_metadata$PAM50_label)
)

analysis_metadata$PAM50_label[
  is.na(analysis_metadata$PAM50_label) |
    analysis_metadata$PAM50_label == ""
] <- "Unclassified"

clean_alteration <- function(alteration_values) {
  toupper(trimws(as.character(alteration_values)))
}

wt_values <- c("NO", "WT", "")
```

# Direct CNV-to-expression associations

The direct gene-level analysis reproduces the data used for Figure 3A. It
compares `PTEN` expression in models with `PTEN` deletion, and `FGFR1` and
`CCND1` expression in models with amplification of the corresponding gene.
Only the target copy-number event and explicitly reported wild-type models are
included. Mutation-only and other non-target events are excluded.

For each comparison, an ordinary least-squares model is fitted with TMM
log2-CPM expression as the dependent variable, CNV status as the main
explanatory variable and Parker PAM50 intrinsic subtype as a covariate. HC3
heteroskedasticity-robust standard errors are used. P values are corrected
across the three comparisons using the Benjamini-Hochberg method. Two-sided
normal-approximation P values are calculated from the robust test statistic,
matching the original Python `statsmodels` analysis.

```{r direct-cnv-expression-definitions}
cnv_expression_definitions <- data.frame(
  gene = c("PTEN", "FGFR1", "CCND1"),
  metadata_column = c(
    "PTEN_alteration",
    "FGFR1_alteration",
    "CCND1_alteration"
  ),
  target_CNV_event = c("DEL", "AMP", "AMP"),
  comparison = c(
    "PTEN deletion",
    "FGFR1 amplification",
    "CCND1 amplification"
  ),
  altered_group = c("Deleted", "Amplified", "Amplified"),
  stringsAsFactors = FALSE
)
```

```{r direct-cnv-expression-associations}
cnv_expression_data_records <- vector(
  mode = "list",
  length = nrow(cnv_expression_definitions)
)

cnv_expression_statistics_records <- vector(
  mode = "list",
  length = nrow(cnv_expression_definitions)
)

for (comparison_index in seq_len(nrow(cnv_expression_definitions))) {

  current_comparison <- cnv_expression_definitions[
    comparison_index,
    ,
    drop = FALSE
  ]

  current_alteration <- clean_alteration(
    analysis_metadata[[current_comparison$metadata_column]]
  )

  current_status <- ifelse(
    current_alteration == current_comparison$target_CNV_event,
    "Altered",
    ifelse(
      current_alteration %in% wt_values,
      "WT",
      NA_character_
    )
  )

  current_expression <- as.numeric(
    expression_log2cpm[
      current_comparison$gene,
      analysis_metadata$samples
    ]
  )

  current_data <- data.frame(
    samples = analysis_metadata$samples,
    gene = current_comparison$gene,
    comparison = current_comparison$comparison,
    target_CNV_event = current_comparison$target_CNV_event,
    altered_group = current_comparison$altered_group,
    reported_alteration = current_alteration,
    CNV_status = current_status,
    PAM50_label = analysis_metadata$PAM50_label,
    expression_TMM_log2CPM = current_expression,
    stringsAsFactors = FALSE
  )

  n_excluded_non_target_events <- sum(
    is.na(current_data$CNV_status)
  )

  current_data <- current_data[
    current_data$CNV_status %in% c("WT", "Altered") &
      complete.cases(
        current_data[, c(
          "expression_TMM_log2CPM",
          "PAM50_label"
        )]
      ),
    ,
    drop = FALSE
  ]

  current_data$CNV_status <- factor(
    current_data$CNV_status,
    levels = c("WT", "Altered")
  )

  current_data$PAM50_label <- factor(
    current_data$PAM50_label
  )

  fitted_model <- lm(
    expression_TMM_log2CPM ~ CNV_status + PAM50_label,
    data = current_data
  )

  robust_coefficients <- lmtest::coeftest(
    fitted_model,
    vcov. = sandwich::vcovHC(
      fitted_model,
      type = "HC3"
    )
  )

  status_coefficient <- robust_coefficients[
    "CNV_statusAltered",
    ,
    drop = FALSE
  ]

  wt_expression <- current_data$expression_TMM_log2CPM[
    current_data$CNV_status == "WT"
  ]

  altered_expression <- current_data$expression_TMM_log2CPM[
    current_data$CNV_status == "Altered"
  ]

  cnv_expression_data_records[[comparison_index]] <- current_data

  cnv_expression_statistics_records[[comparison_index]] <- data.frame(
    gene = current_comparison$gene,
    comparison = current_comparison$comparison,
    target_CNV_event = current_comparison$target_CNV_event,
    altered_group = current_comparison$altered_group,
    n_WT = length(wt_expression),
    n_altered = length(altered_expression),
    n_excluded_non_target_events = n_excluded_non_target_events,
    mean_WT_log2CPM = mean(wt_expression),
    mean_altered_log2CPM = mean(altered_expression),
    median_WT_log2CPM = median(wt_expression),
    median_altered_log2CPM = median(altered_expression),
    beta_altered_vs_WT_log2CPM = status_coefficient[1, "Estimate"],
    robust_standard_error = status_coefficient[1, "Std. Error"],
    robust_z_statistic = status_coefficient[1, "t value"],
    p_value = 2 * pnorm(
      -abs(status_coefficient[1, "t value"])
    ),
    adjustments = "PAM50 intrinsic subtype",
    stringsAsFactors = FALSE
  )
}

direct_cnv_expression_data <- do.call(
  rbind,
  cnv_expression_data_records
)

direct_cnv_expression_statistics <- do.call(
  rbind,
  cnv_expression_statistics_records
)

direct_cnv_expression_statistics$FDR_q_global <- p.adjust(
  direct_cnv_expression_statistics$p_value,
  method = "BH"
)

write.csv(
  direct_cnv_expression_data,
  file.path(
    results_dir,
    paste0(
      "Direct_CNV_to_expression_associations_",
      "PAM50_adjusted_model_data.csv"
    )
  ),
  row.names = FALSE,
  quote = TRUE
)

write.csv(
  direct_cnv_expression_statistics,
  file.path(
    results_dir,
    paste0(
      "Direct_CNV_to_expression_associations_",
      "PAM50_adjusted_statistics.csv"
    )
  ),
  row.names = FALSE,
  quote = TRUE
)

direct_cnv_expression_statistics
```

# PAM50-adjusted genomic-to-transcriptomic associations

The following analyses reproduce the alteration definitions and statistical
comparisons used to relate genomic features to Hallmark pathway activity. Each
model uses the ssGSEA score as the dependent variable, genomic status as the
main explanatory variable and Parker PAM50 intrinsic subtype as a covariate.
HC3 heteroskedasticity-robust standard errors and two-sided normal-approximation
P values are used, matching the original Python analyses. Mann-Whitney tests
are retained as unadjusted descriptive comparisons.

```{r pathway-association-functions}
hallmark_score <- function(pathway_name) {
  as.numeric(
    hallmark_ssgsea_scores[
      pathway_name,
      analysis_metadata$samples
    ]
  )
}

is_any_alteration <- function(alteration_values) {
  cleaned_values <- clean_alteration(alteration_values)
  !is.na(cleaned_values) &
    !cleaned_values %in% c("", "NO", "WT", "MET", "NAN", "NONE")
}

observed_alteration_types <- function(alteration_values) {
  cleaned_values <- clean_alteration(alteration_values)
  altered_values <- cleaned_values[
    is_any_alteration(cleaned_values)
  ]
  paste(sort(unique(altered_values)), collapse = ", ")
}

altered_gene_labels <- function(alteration_matrix) {
  apply(
    alteration_matrix,
    1,
    function(current_row) {
      paste(names(current_row)[as.logical(current_row)], collapse = ", ")
    }
  )
}

fit_pam50_adjusted_model <- function(
    score,
    status,
    reference_label,
    altered_label,
    pam50_subtype = analysis_metadata$PAM50_label) {

  model_data <- data.frame(
    Score = score,
    Status = status,
    PAM50_subtype = pam50_subtype,
    stringsAsFactors = FALSE
  )

  model_data <- model_data[
    complete.cases(model_data),
    ,
    drop = FALSE
  ]

  model_data$Status <- factor(
    model_data$Status,
    levels = c(reference_label, altered_label)
  )

  model_data$PAM50_subtype <- factor(
    model_data$PAM50_subtype
  )

  fitted_model <- lm(
    Score ~ Status + PAM50_subtype,
    data = model_data
  )

  robust_covariance <- sandwich::vcovHC(
    fitted_model,
    type = "HC3"
  )

  status_term <- grep(
    "^Status",
    names(coef(fitted_model)),
    value = TRUE
  )[1]

  beta <- unname(coef(fitted_model)[status_term])
  robust_standard_error <- sqrt(
    robust_covariance[status_term, status_term]
  )
  robust_z_statistic <- beta / robust_standard_error

  reference_scores <- model_data$Score[
    model_data$Status == reference_label
  ]

  altered_scores <- model_data$Score[
    model_data$Status == altered_label
  ]

  list(
    n_reference = length(reference_scores),
    n_altered = length(altered_scores),
    mean_reference = mean(reference_scores),
    mean_altered = mean(altered_scores),
    median_reference = median(reference_scores),
    median_altered = median(altered_scores),
    difference_raw_means = mean(altered_scores) - mean(reference_scores),
    beta = beta,
    CI95_lower = beta - qnorm(0.975) * robust_standard_error,
    CI95_upper = beta + qnorm(0.975) * robust_standard_error,
    p_value_HC3 = 2 * pnorm(-abs(robust_z_statistic)),
    p_value_Mann_Whitney = wilcox.test(
      altered_scores,
      reference_scores,
      alternative = "two.sided"
    )$p.value,
    model_R_squared = summary(fitted_model)$r.squared,
    reference_scores = reference_scores,
    altered_scores = altered_scores
  )
}
```

# Estrogen-response pathway alterations

The estrogen-pathway analysis defines any alteration in `ESR1`, `GATA3`,
`FOXA1`, `KMT2C` or `CCND1` as altered, while `NO`, `WT`, missing values and
`MET` are treated as no alteration. Composite and individual-gene associations
are tested against the mean of the early and late estrogen-response scores and
against each score separately.

```{r estrogen-pathway-status}
estrogen_genes <- c("ESR1", "GATA3", "FOXA1", "KMT2C", "CCND1")
estrogen_columns <- paste0(estrogen_genes, "_alteration")

estrogen_status <- data.frame(
  samples = analysis_metadata$samples,
  HALLMARK_ESTROGEN_RESPONSE_EARLY = hallmark_score(
    "HALLMARK_ESTROGEN_RESPONSE_EARLY"
  ),
  HALLMARK_ESTROGEN_RESPONSE_LATE = hallmark_score(
    "HALLMARK_ESTROGEN_RESPONSE_LATE"
  ),
  analysis_metadata[, estrogen_columns, drop = FALSE],
  PAM50_subtype = analysis_metadata$PAM50_label,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

estrogen_status$Mean_estrogen_response <- rowMeans(
  estrogen_status[, c(
    "HALLMARK_ESTROGEN_RESPONSE_EARLY",
    "HALLMARK_ESTROGEN_RESPONSE_LATE"
  )]
)

estrogen_alteration_matrix <- sapply(
  estrogen_columns,
  function(current_column) {
    is_any_alteration(estrogen_status[[current_column]])
  }
)
colnames(estrogen_alteration_matrix) <- estrogen_genes

for (gene_name in estrogen_genes) {
  estrogen_status[[paste0(gene_name, "_altered")]] <-
    estrogen_alteration_matrix[, gene_name]
}

estrogen_status$Altered_estrogen_pathway_genes <- altered_gene_labels(
  estrogen_alteration_matrix
)

estrogen_status$N_estrogen_pathway_gene_alterations <- rowSums(
  estrogen_alteration_matrix
)

estrogen_status$Estrogen_pathway_status <- ifelse(
  estrogen_status$N_estrogen_pathway_gene_alterations > 0,
  ">=1 alteration",
  "No alteration"
)

write.csv(
  estrogen_status,
  file.path(
    results_dir,
    "Composite_estrogen_pathway_alteration_PDX_status.csv"
  ),
  row.names = FALSE,
  quote = TRUE
)
```

```{r composite-estrogen-statistics}
estrogen_outcomes <- data.frame(
  outcome = c(
    "Mean Estrogen Response",
    "Estrogen Response Early",
    "Estrogen Response Late"
  ),
  score_column = c(
    "Mean_estrogen_response",
    "HALLMARK_ESTROGEN_RESPONSE_EARLY",
    "HALLMARK_ESTROGEN_RESPONSE_LATE"
  ),
  stringsAsFactors = FALSE
)

composite_estrogen_records <- vector(
  mode = "list",
  length = nrow(estrogen_outcomes)
)

for (outcome_index in seq_len(nrow(estrogen_outcomes))) {
  current_outcome <- estrogen_outcomes[outcome_index, , drop = FALSE]

  current_model <- fit_pam50_adjusted_model(
    score = estrogen_status[[current_outcome$score_column]],
    status = estrogen_status$Estrogen_pathway_status,
    reference_label = "No alteration",
    altered_label = ">=1 alteration"
  )

  composite_estrogen_records[[outcome_index]] <- data.frame(
    outcome = current_outcome$outcome,
    n_no_alteration = current_model$n_reference,
    n_at_least_one_alteration = current_model$n_altered,
    mean_no_alteration = current_model$mean_reference,
    mean_at_least_one_alteration = current_model$mean_altered,
    difference_raw_means = current_model$difference_raw_means,
    adjustments = "PAM50 intrinsic subtype",
    beta_at_least_one_alteration_vs_none = current_model$beta,
    CI95_lower = current_model$CI95_lower,
    CI95_upper = current_model$CI95_upper,
    p_value_PAM50_adjusted_HC3 = current_model$p_value_HC3,
    p_value_Mann_Whitney_unadjusted_descriptive =
      current_model$p_value_Mann_Whitney,
    model_R_squared = current_model$model_R_squared,
    stringsAsFactors = FALSE
  )
}

composite_estrogen_statistics <- do.call(
  rbind,
  composite_estrogen_records
)

composite_estrogen_statistics$FDR_q_across_3_estrogen_scores <- p.adjust(
  composite_estrogen_statistics$p_value_PAM50_adjusted_HC3,
  method = "BH"
)

write.csv(
  composite_estrogen_statistics,
  file.path(
    results_dir,
    paste0(
      "Composite_estrogen_pathway_alteration_",
      "PAM50_adjusted_statistics.csv"
    )
  ),
  row.names = FALSE,
  quote = TRUE
)
```

```{r individual-estrogen-statistics}
individual_estrogen_records <- list()
record_index <- 1

for (gene_index in seq_along(estrogen_genes)) {
  gene_name <- estrogen_genes[gene_index]
  alteration_column <- estrogen_columns[gene_index]
  altered_flag <- estrogen_alteration_matrix[, gene_name]
  current_status <- ifelse(altered_flag, "Altered", "WT")
  observed_types <- observed_alteration_types(
    estrogen_status[[alteration_column]]
  )

  for (outcome_index in seq_len(nrow(estrogen_outcomes))) {
    current_outcome <- estrogen_outcomes[outcome_index, , drop = FALSE]

    current_model <- fit_pam50_adjusted_model(
      score = estrogen_status[[current_outcome$score_column]],
      status = current_status,
      reference_label = "WT",
      altered_label = "Altered"
    )

    individual_estrogen_records[[record_index]] <- data.frame(
      feature = paste(gene_name, "alteration"),
      metadata_column = alteration_column,
      altered_definition = "Any alteration; MET treated as WT",
      observed_alteration_types = observed_types,
      outcome = current_outcome$outcome,
      n_WT = current_model$n_reference,
      n_altered = current_model$n_altered,
      mean_WT = current_model$mean_reference,
      mean_altered = current_model$mean_altered,
      difference_raw_means = current_model$difference_raw_means,
      adjustments = "PAM50 intrinsic subtype",
      model_status = "success",
      interpretation_note = ifelse(
        current_model$n_altered < 5,
        paste0(
          "Exploratory: fewer than 5 altered PDXs"
        ),
        ""
      ),
      beta_altered_vs_WT_PAM50_adjusted = current_model$beta,
      CI95_lower = current_model$CI95_lower,
      CI95_upper = current_model$CI95_upper,
      p_value_PAM50_adjusted_HC3 = current_model$p_value_HC3,
      p_value_Mann_Whitney_unadjusted_descriptive =
        current_model$p_value_Mann_Whitney,
      model_R_squared = current_model$model_R_squared,
      stringsAsFactors = FALSE
    )

    record_index <- record_index + 1
  }
}

individual_estrogen_statistics <- do.call(
  rbind,
  individual_estrogen_records
)

individual_estrogen_statistics$FDR_q_global_15_tests <- p.adjust(
  individual_estrogen_statistics$p_value_PAM50_adjusted_HC3,
  method = "BH"
)

individual_estrogen_statistics <- individual_estrogen_statistics[
  order(
    individual_estrogen_statistics$FDR_q_global_15_tests,
    individual_estrogen_statistics$p_value_PAM50_adjusted_HC3
  ),
  ,
  drop = FALSE
]

write.csv(
  individual_estrogen_statistics,
  file.path(
    results_dir,
    paste0(
      "Estrogen_response_ESR1_pathway_alterations_",
      "PAM50_adjusted_statistics.csv"
    )
  ),
  row.names = FALSE,
  quote = TRUE
)
```

# PI3K/AKT/mTOR-pathway alterations

The PI3K/AKT/mTOR analysis evaluates a composite alteration status across
`PIK3CA`, `AKT1`, `PTEN` and `TSC1`, as well as each gene separately. Any
non-reference event is considered altered, except `MET`, which is treated as
wild type. Models with fewer than three altered PDXs are reported but are not
formally tested.

```{r pi3k-pathway-status}
pi3k_genes <- c("PIK3CA", "AKT1", "PTEN", "TSC1")
pi3k_columns <- paste0(pi3k_genes, "_alteration")

pi3k_status <- data.frame(
  samples = analysis_metadata$samples,
  HALLMARK_PI3K_AKT_MTOR_SIGNALING = hallmark_score(
    "HALLMARK_PI3K_AKT_MTOR_SIGNALING"
  ),
  HALLMARK_MTORC1_SIGNALING = hallmark_score(
    "HALLMARK_MTORC1_SIGNALING"
  ),
  analysis_metadata[, pi3k_columns, drop = FALSE],
  PAM50_subtype = analysis_metadata$PAM50_label,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

pi3k_status$Mean_PI3K_AKT_mTOR_activity <- rowMeans(
  pi3k_status[, c(
    "HALLMARK_PI3K_AKT_MTOR_SIGNALING",
    "HALLMARK_MTORC1_SIGNALING"
  )]
)

pi3k_alteration_matrix <- sapply(
  pi3k_columns,
  function(current_column) {
    is_any_alteration(pi3k_status[[current_column]])
  }
)
colnames(pi3k_alteration_matrix) <- pi3k_genes

for (gene_name in pi3k_genes) {
  pi3k_status[[paste0(gene_name, "_altered")]] <-
    pi3k_alteration_matrix[, gene_name]
}

pi3k_status$Altered_PI3K_AKT_mTOR_genes <- altered_gene_labels(
  pi3k_alteration_matrix
)

pi3k_status$N_PI3K_AKT_mTOR_gene_alterations <- rowSums(
  pi3k_alteration_matrix
)

pi3k_status$Composite_PI3K_AKT_mTOR_status <- ifelse(
  pi3k_status$N_PI3K_AKT_mTOR_gene_alterations > 0,
  ">=1 alteration",
  "No alteration"
)

write.csv(
  pi3k_status,
  file.path(
    results_dir,
    "PI3K_AKT_mTOR_alterations_PDX_status.csv"
  ),
  row.names = FALSE,
  quote = TRUE
)
```

```{r pi3k-pathway-statistics}
pi3k_outcomes <- data.frame(
  outcome = c(
    "Mean PI3K/AKT/mTOR activity",
    "PI3K/AKT/mTOR signaling",
    "mTORC1 signaling"
  ),
  score_column = c(
    "Mean_PI3K_AKT_mTOR_activity",
    "HALLMARK_PI3K_AKT_MTOR_SIGNALING",
    "HALLMARK_MTORC1_SIGNALING"
  ),
  stringsAsFactors = FALSE
)

pi3k_comparisons <- list(
  list(
    feature_type = "Composite",
    feature = ">=1 PI3K/AKT/mTOR alteration",
    status = pi3k_status$Composite_PI3K_AKT_mTOR_status,
    reference_label = "No alteration",
    altered_label = ">=1 alteration",
    alteration_definition = paste0(
      "At least one alteration in PIK3CA, AKT1, PTEN, or TSC1; ",
      "MET treated as WT"
    ),
    observed_types = "Composite of PIK3CA, AKT1, PTEN, and TSC1 events"
  )
)

for (gene_index in seq_along(pi3k_genes)) {
  gene_name <- pi3k_genes[gene_index]
  alteration_column <- pi3k_columns[gene_index]

  pi3k_comparisons[[length(pi3k_comparisons) + 1]] <- list(
    feature_type = "Individual gene",
    feature = paste(gene_name, "alteration"),
    status = ifelse(
      pi3k_alteration_matrix[, gene_name],
      "Altered",
      "WT"
    ),
    reference_label = "WT",
    altered_label = "Altered",
    alteration_definition = "Any alteration; MET treated as WT",
    observed_types = observed_alteration_types(
      pi3k_status[[alteration_column]]
    )
  )
}

pi3k_statistics_records <- list()
record_index <- 1

for (current_comparison in pi3k_comparisons) {
  for (outcome_index in seq_len(nrow(pi3k_outcomes))) {
    current_outcome <- pi3k_outcomes[outcome_index, , drop = FALSE]
    n_altered <- sum(
      current_comparison$status == current_comparison$altered_label
    )
    n_reference <- sum(
      current_comparison$status == current_comparison$reference_label
    )
    current_score <- pi3k_status[[current_outcome$score_column]]
    reference_scores <- current_score[
      current_comparison$status == current_comparison$reference_label
    ]
    altered_scores <- current_score[
      current_comparison$status == current_comparison$altered_label
    ]

    current_record <- data.frame(
      feature_type = current_comparison$feature_type,
      feature = current_comparison$feature,
      alteration_definition = current_comparison$alteration_definition,
      observed_alteration_types = current_comparison$observed_types,
      outcome = current_outcome$outcome,
      n_WT = n_reference,
      n_altered = n_altered,
      mean_WT = mean(reference_scores),
      mean_altered = mean(altered_scores),
      difference_raw_means = mean(altered_scores) - mean(reference_scores),
      adjustments = "PAM50 intrinsic subtype",
      model_status = ifelse(
        n_altered < 3,
        "not_tested_too_few_altered_PDXs",
        "success"
      ),
      interpretation_note = ifelse(
        n_altered < 3,
        paste0(
          "Only ", n_altered,
          " altered PDX(s); formal model not calculated."
        ),
        ifelse(
          n_altered < 5,
          "Exploratory: fewer than five altered PDXs.",
          ""
        )
      ),
      beta_altered_vs_WT_PAM50_adjusted = NA_real_,
      CI95_lower = NA_real_,
      CI95_upper = NA_real_,
      p_value_PAM50_adjusted_HC3 = NA_real_,
      p_value_Mann_Whitney_unadjusted_descriptive = NA_real_,
      model_R_squared = NA_real_,
      stringsAsFactors = FALSE
    )

    if (n_altered >= 3) {
      current_model <- fit_pam50_adjusted_model(
        score = current_score,
        status = current_comparison$status,
        reference_label = current_comparison$reference_label,
        altered_label = current_comparison$altered_label
      )

      current_record$beta_altered_vs_WT_PAM50_adjusted <-
        current_model$beta
      current_record$CI95_lower <- current_model$CI95_lower
      current_record$CI95_upper <- current_model$CI95_upper
      current_record$p_value_PAM50_adjusted_HC3 <-
        current_model$p_value_HC3
      current_record$p_value_Mann_Whitney_unadjusted_descriptive <-
        current_model$p_value_Mann_Whitney
      current_record$model_R_squared <- current_model$model_R_squared
    }

    pi3k_statistics_records[[record_index]] <- current_record
    record_index <- record_index + 1
  }
}

pi3k_statistics <- do.call(rbind, pi3k_statistics_records)
pi3k_statistics$FDR_q_global_all_valid_tests <- NA_real_

valid_pi3k_tests <- pi3k_statistics$model_status == "success"

pi3k_statistics$FDR_q_global_all_valid_tests[valid_pi3k_tests] <- p.adjust(
  pi3k_statistics$p_value_PAM50_adjusted_HC3[valid_pi3k_tests],
  method = "BH"
)

pi3k_statistics <- pi3k_statistics[
  order(
    pi3k_statistics$FDR_q_global_all_valid_tests,
    pi3k_statistics$p_value_PAM50_adjusted_HC3,
    na.last = TRUE
  ),
  ,
  drop = FALSE
]

write.csv(
  pi3k_statistics,
  file.path(
    results_dir,
    "PI3K_AKT_mTOR_alterations_PAM50_adjusted_statistics.csv"
  ),
  row.names = FALSE,
  quote = TRUE
)
```

# Strict PTEN-deletion analysis

This analysis compares `PTEN` deletion with explicitly reported `NO` calls.
Other `PTEN` events are excluded. The two displayed pathway signatures are
corrected together.

```{r strict-pten-deletion-statistics}
strict_pten_calls <- clean_alteration(
  analysis_metadata$PTEN_alteration
)
strict_pten_calls[is.na(strict_pten_calls)] <- "NO"

strict_pten_keep <- strict_pten_calls %in% c("NO", "DEL")
strict_pten_group <- ifelse(
  strict_pten_calls[strict_pten_keep] == "DEL",
  "PTEN deletion",
  "PTEN WT"
)

strict_pten_outcomes <- data.frame(
  outcome = c(
    "PI3K/AKT/mTOR signaling",
    "mTORC1 signaling"
  ),
  pathway = c(
    "HALLMARK_PI3K_AKT_MTOR_SIGNALING",
    "HALLMARK_MTORC1_SIGNALING"
  ),
  stringsAsFactors = FALSE
)

strict_pten_records <- vector(
  mode = "list",
  length = nrow(strict_pten_outcomes)
)

for (outcome_index in seq_len(nrow(strict_pten_outcomes))) {
  current_outcome <- strict_pten_outcomes[outcome_index, , drop = FALSE]
  current_score <- hallmark_score(current_outcome$pathway)[strict_pten_keep]

  current_model <- fit_pam50_adjusted_model(
    score = current_score,
    status = strict_pten_group,
    reference_label = "PTEN WT",
    altered_label = "PTEN deletion",
    pam50_subtype = analysis_metadata$PAM50_label[strict_pten_keep]
  )

  strict_pten_records[[outcome_index]] <- data.frame(
    outcome = current_outcome$outcome,
    n_PTEN_WT = current_model$n_reference,
    n_PTEN_deletion = current_model$n_altered,
    mean_PTEN_WT = current_model$mean_reference,
    mean_PTEN_deletion = current_model$mean_altered,
    median_PTEN_WT = current_model$median_reference,
    median_PTEN_deletion = current_model$median_altered,
    difference_raw_means = current_model$difference_raw_means,
    adjustments = "PAM50 intrinsic subtype",
    beta_PTEN_deletion_vs_WT = current_model$beta,
    CI95_lower = current_model$CI95_lower,
    CI95_upper = current_model$CI95_upper,
    p_value_PAM50_adjusted_HC3 = current_model$p_value_HC3,
    p_value_Mann_Whitney_unadjusted =
      current_model$p_value_Mann_Whitney,
    p_value_Welch_t_test_unadjusted = t.test(
      current_model$altered_scores,
      current_model$reference_scores,
      var.equal = FALSE
    )$p.value,
    model_R_squared = current_model$model_R_squared,
    stringsAsFactors = FALSE
  )
}

strict_pten_statistics <- do.call(rbind, strict_pten_records)

strict_pten_statistics$FDR_q_across_2_signatures <- p.adjust(
  strict_pten_statistics$p_value_PAM50_adjusted_HC3,
  method = "BH"
)

write.csv(
  strict_pten_statistics,
  file.path(
    results_dir,
    "PTEN_deletion_PI3K_mTORC1_PAM50_adjusted_statistics.csv"
  ),
  row.names = FALSE,
  quote = TRUE
)
```

# RB/cell-cycle dysregulation

RB/cell-cycle dysregulation is defined as `RB1` functional loss (`DEL` or
`TRU`), `CDKN2B` deletion, or `CCND1` amplification. Composite and
individual-gene associations are tested against E2F-target, G2/M-checkpoint
and mitotic-spindle activity.

```{r rb-cell-cycle-status}
rb1_calls <- clean_alteration(analysis_metadata$RB1_alteration)
cdkn2b_calls <- clean_alteration(analysis_metadata$CDKN2B_alteration)
ccnd1_calls <- clean_alteration(analysis_metadata$CCND1_alteration)

rb1_loss <- rb1_calls %in% c("DEL", "TRU")
cdkn2b_loss <- !is.na(cdkn2b_calls) & cdkn2b_calls == "DEL"
ccnd1_amplification <- !is.na(ccnd1_calls) & ccnd1_calls == "AMP"

cell_cycle_event_matrix <- cbind(
  RB1 = rb1_loss,
  CDKN2B = cdkn2b_loss,
  CCND1 = ccnd1_amplification
)

rb_cell_cycle_status <- data.frame(
  samples = analysis_metadata$samples,
  HALLMARK_E2F_TARGETS = hallmark_score("HALLMARK_E2F_TARGETS"),
  HALLMARK_G2M_CHECKPOINT = hallmark_score("HALLMARK_G2M_CHECKPOINT"),
  HALLMARK_MITOTIC_SPINDLE = hallmark_score("HALLMARK_MITOTIC_SPINDLE"),
  RB1_alteration = analysis_metadata$RB1_alteration,
  CDKN2B_alteration = analysis_metadata$CDKN2B_alteration,
  CCND1_alteration = analysis_metadata$CCND1_alteration,
  PAM50_subtype = analysis_metadata$PAM50_label,
  RB1_loss = rb1_loss,
  CDKN2B_loss = cdkn2b_loss,
  CCND1_amplification = ccnd1_amplification,
  Composite_cell_cycle_dysregulation = rowSums(cell_cycle_event_matrix) > 0,
  stringsAsFactors = FALSE
)

rb_cell_cycle_status$Altered_cell_cycle_genes <- altered_gene_labels(
  cell_cycle_event_matrix
)

rb_cell_cycle_status$N_cell_cycle_events <- rowSums(
  cell_cycle_event_matrix
)

rb_cell_cycle_status$Composite_cell_cycle_status <- ifelse(
  rb_cell_cycle_status$Composite_cell_cycle_dysregulation,
  "Altered",
  "WT"
)

write.csv(
  rb_cell_cycle_status,
  file.path(
    results_dir,
    "RB_cell_cycle_dysregulation_PDX_status.csv"
  ),
  row.names = FALSE,
  quote = TRUE
)
```

```{r rb-cell-cycle-statistics}
cell_cycle_outcomes <- data.frame(
  outcome = c("E2F Targets", "G2M Checkpoint", "Mitotic Spindle"),
  score_column = c(
    "HALLMARK_E2F_TARGETS",
    "HALLMARK_G2M_CHECKPOINT",
    "HALLMARK_MITOTIC_SPINDLE"
  ),
  stringsAsFactors = FALSE
)

cell_cycle_comparisons <- list(
  list(
    feature_type = "Composite",
    feature = "RB/cell-cycle dysregulation",
    altered_definition = paste0(
      "RB1 functional loss, CDKN2B functional loss, ",
      "or CCND1 amplification"
    ),
    observed_types = "RB1: DEL, TRU | CDKN2B: DEL | CCND1: AMP",
    status = rb_cell_cycle_status$Composite_cell_cycle_status
  ),
  list(
    feature_type = "Individual gene",
    feature = "RB1 loss",
    altered_definition = "Functional loss event",
    observed_types = "DEL, TRU",
    status = ifelse(rb1_loss, "Altered", "WT")
  ),
  list(
    feature_type = "Individual gene",
    feature = "CDKN2B loss",
    altered_definition = "Deletion only",
    observed_types = "DEL",
    status = ifelse(cdkn2b_loss, "Altered", "WT")
  ),
  list(
    feature_type = "Individual gene",
    feature = "CCND1 amplification",
    altered_definition = "Amplification only",
    observed_types = "AMP",
    status = ifelse(ccnd1_amplification, "Altered", "WT")
  )
)

cell_cycle_statistics_records <- list()
record_index <- 1

for (current_comparison in cell_cycle_comparisons) {
  for (outcome_index in seq_len(nrow(cell_cycle_outcomes))) {
    current_outcome <- cell_cycle_outcomes[outcome_index, , drop = FALSE]

    current_model <- fit_pam50_adjusted_model(
      score = rb_cell_cycle_status[[current_outcome$score_column]],
      status = current_comparison$status,
      reference_label = "WT",
      altered_label = "Altered"
    )

    cell_cycle_statistics_records[[record_index]] <- data.frame(
      feature_type = current_comparison$feature_type,
      feature = current_comparison$feature,
      altered_definition = current_comparison$altered_definition,
      observed_alteration_types = current_comparison$observed_types,
      outcome = current_outcome$outcome,
      n_WT = current_model$n_reference,
      n_altered = current_model$n_altered,
      mean_WT = current_model$mean_reference,
      mean_altered = current_model$mean_altered,
      difference_raw_means = current_model$difference_raw_means,
      adjustments = "PAM50 intrinsic subtype",
      model_status = "success",
      interpretation_note = "",
      beta_altered_vs_WT_PAM50_adjusted = current_model$beta,
      CI95_lower = current_model$CI95_lower,
      CI95_upper = current_model$CI95_upper,
      p_value_PAM50_adjusted_HC3 = current_model$p_value_HC3,
      p_value_Mann_Whitney_unadjusted_descriptive =
        current_model$p_value_Mann_Whitney,
      model_R_squared = current_model$model_R_squared,
      stringsAsFactors = FALSE
    )

    record_index <- record_index + 1
  }
}

rb_cell_cycle_statistics <- do.call(
  rbind,
  cell_cycle_statistics_records
)

rb_cell_cycle_statistics$FDR_q_global_all_valid_tests <- p.adjust(
  rb_cell_cycle_statistics$p_value_PAM50_adjusted_HC3,
  method = "BH"
)

rb_cell_cycle_statistics <- rb_cell_cycle_statistics[
  order(
    rb_cell_cycle_statistics$FDR_q_global_all_valid_tests,
    rb_cell_cycle_statistics$p_value_PAM50_adjusted_HC3
  ),
  ,
  drop = FALSE
]

write.csv(
  rb_cell_cycle_statistics,
  file.path(
    results_dir,
    "RB_cell_cycle_dysregulation_PAM50_adjusted_statistics.csv"
  ),
  row.names = FALSE,
  quote = TRUE
)

list(
  composite_estrogen = composite_estrogen_statistics,
  individual_estrogen = individual_estrogen_statistics,
  pi3k_akt_mtor = pi3k_statistics,
  strict_pten_deletion = strict_pten_statistics,
  rb_cell_cycle = rb_cell_cycle_statistics
)
```

# Reproducibility information

```{r session-information}
sessionInfo()
```
