---
title: "Baseline Hallmark GSEA across drug-response cohorts"
author: "Andreu Odena"
date: "2026-09-08"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Objective

This analysis identifies baseline Hallmark transcriptional programmes
associated with sensitivity and resistance across ten treatment cohorts.
Differential-expression statistics are adjusted for Parker PAM50 intrinsic
subtype when this covariate is estimable and are used to rank genes for GSEA.
Positive enrichment scores indicate higher pathway activity in sensitive
models, whereas negative scores indicate higher activity in resistant models.

No figures are generated by this R Markdown.

# Packages and paths

The analysis should be run from the root of the GitHub repository. Raw counts
and sensitivity annotations are read from the local `data` directory. PAM50
calls are generated by `02_PAM50_classification.Rmd`.

```{r packages-and-paths}
library(edgeR)
library(limma)
library(fgsea)
library(msigdbr)

counts_path <- file.path("data", "260824_Counts.csv")
metadata_path <- file.path("data", "260824_Metadata.csv")
pam50_path <- file.path(
  "results",
  "PAM50",
  "PAM50_Parker_calls.csv"
)

results_dir <- file.path(
  "results",
  "drug_response_Hallmark_GSEA"
)

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

set.seed(260824)
```

# Load and normalize RNA-seq data

Genes are retained when they have more than 10 counts in at least 15 PDX
models. Library-size and composition differences are normalized using TMM,
followed by transformation to log2 counts per million.

```{r load-and-normalize-expression}
counts_input <- read.csv(
  counts_path,
  sep = ";",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

metadata <- read.csv(
  metadata_path,
  sep = ";",
  check.names = FALSE,
  stringsAsFactors = FALSE,
  fileEncoding = "UTF-8-BOM"
)

pam50_calls <- read.csv(
  pam50_path,
  check.names = FALSE,
  stringsAsFactors = FALSE,
  fileEncoding = "UTF-8-BOM"
)

gene_names <- toupper(trimws(as.character(counts_input$gene_name)))
counts_matrix <- as.matrix(counts_input[, -1, drop = FALSE])
storage.mode(counts_matrix) <- "numeric"
rownames(counts_matrix) <- gene_names

metadata$samples <- trimws(as.character(metadata$samples))
pam50_calls$samples <- trimws(as.character(pam50_calls$samples))

metadata <- metadata[
  match(colnames(counts_matrix), metadata$samples),
  ,
  drop = FALSE
]

keep_gene <- rowSums(counts_matrix > 10) >= 15
counts_filtered <- counts_matrix[keep_gene, , drop = FALSE]

dge <- DGEList(counts = counts_filtered)
dge <- calcNormFactors(dge, method = "TMM")

expression_log2cpm <- cpm(
  dge,
  log = TRUE,
  prior.count = 1
)

expression_log2cpm <- limma::avereps(
  expression_log2cpm,
  ID = rownames(expression_log2cpm)
)
```

# Define the treatment-response cohorts

Curated `Sensitive` and `Resistant` annotations are used directly. For columns
containing mRECIST categories, progressive disease is classified as resistant,
whereas stable disease, partial response and complete response are classified
as sensitive.

```{r response-definitions}
normalize_response <- function(response_values) {
  cleaned_values <- toupper(trimws(as.character(response_values)))
  cleaned_values[cleaned_values %in% c(
    "", "NA", "N/A", "NOT TESTED", "NONE"
  )] <- NA_character_

  response_group <- rep(NA_character_, length(cleaned_values))

  response_group[cleaned_values %in% c(
    "RESISTANT", "RESISTANCE", "PD", "PROGRESSIVE DISEASE"
  )] <- "Resistant"

  response_group[cleaned_values %in% c(
    "SENSITIVE", "SENSTIVE", "SENSITIVITY",
    "SD", "PR", "CR", "SD/PR/CR",
    "STABLE DISEASE", "PARTIAL RESPONSE", "COMPLETE RESPONSE"
  )] <- "Sensitive"

  response_group
}

treatment_definitions <- data.frame(
  treatment = c(
    "PARP inhibitor",
    "Platinum chemotherapy",
    "Irinotecan (TOP1i)",
    "Adavosertib (WEE1i)",
    "PI3K inhibitor",
    "CDK4/6 inhibitor",
    "SERD",
    "AKT inhibitor",
    "Antimicrotubule chemotherapy",
    "FGFR inhibitor"
  ),
  metadata_column = c(
    "PARPi_sensitivity",
    "Plat_chemo_sensitivity",
    "Irinotecan_(TOP1i_chemo)_response",
    "Adavosertib_(AZD1775)_(WEE1i)_response",
    "PI3Ki_sensitivity",
    "CDK4-6i_sensitivity",
    "SERD_sensitivity",
    "AKTi_sensitivity",
    "antimicrotubule_chemo_sensitivity",
    "FGFRi_sensitivity"
  ),
  stringsAsFactors = FALSE
)
```

# Hallmark gene sets

Hallmark gene sets are obtained from MSigDB through `msigdbr`. Gene symbols are
converted to uppercase and gene sets containing fewer than 10 genes are
excluded.

```{r hallmark-gene-sets}
hallmark_table <- msigdbr::msigdbr(
  species = "Homo sapiens",
  collection = "H"
)

hallmark_pathways <- split(
  toupper(hallmark_table$gene_symbol),
  hallmark_table$gs_name
)

hallmark_pathways <- lapply(
  hallmark_pathways,
  unique
)

hallmark_pathways <- hallmark_pathways[
  lengths(hallmark_pathways) >= 10
]
```

# PAM50-adjusted differential expression and GSEA

Within each treatment cohort, gene-level expression is modelled using limma.
The response coefficient compares sensitive with resistant models and PAM50
intrinsic subtype is included as a covariate whenever at least two subtypes are
represented. For cohorts containing a single subtype, only response group is
included. Moderated *t*-statistics are ranked and analyzed using
`fgseaMultilevel`.

```{r treatment-gsea-function}
run_treatment_gsea <- function(
    treatment_name,
    response_column,
    metadata,
    pam50_calls,
    expression_matrix,
    hallmark_pathways) {

  analysis_table <- data.frame(
    samples = metadata$samples,
    raw_response = metadata[[response_column]],
    response_group = normalize_response(metadata[[response_column]]),
    stringsAsFactors = FALSE
  )

  analysis_table <- merge(
    analysis_table,
    pam50_calls[, c("samples", "PAM50_label")],
    by = "samples",
    all.x = TRUE,
    sort = FALSE
  )

  colnames(analysis_table)[
    colnames(analysis_table) == "PAM50_label"
  ] <- "PAM50_subtype"

  analysis_table$PAM50_subtype <- trimws(
    as.character(analysis_table$PAM50_subtype)
  )

  analysis_table <- analysis_table[
    analysis_table$samples %in% colnames(expression_matrix) &
      !is.na(analysis_table$response_group) &
      !is.na(analysis_table$PAM50_subtype) &
      analysis_table$PAM50_subtype != "",
    ,
    drop = FALSE
  ]

  selected_samples <- colnames(expression_matrix)[
    colnames(expression_matrix) %in% analysis_table$samples
  ]

  analysis_table <- analysis_table[
    match(selected_samples, analysis_table$samples),
    ,
    drop = FALSE
  ]

  analysis_table$response_group <- factor(
    analysis_table$response_group,
    levels = c("Resistant", "Sensitive")
  )

  n_resistant <- sum(
    analysis_table$response_group == "Resistant"
  )
  n_sensitive <- sum(
    analysis_table$response_group == "Sensitive"
  )
  n_total <- nrow(analysis_table)
  n_pam50_subtypes <- length(unique(analysis_table$PAM50_subtype))

  if (n_pam50_subtypes >= 2) {
    analysis_table$PAM50_subtype <- factor(
      analysis_table$PAM50_subtype
    )

    design_matrix <- model.matrix(
      ~ PAM50_subtype + response_group,
      data = analysis_table
    )

    adjustment_label <- "PAM50 intrinsic subtype"
    model_note <- "PAM50 intrinsic subtype included as a covariate."
  } else {
    design_matrix <- model.matrix(
      ~ response_group,
      data = analysis_table
    )

    adjustment_label <- paste0(
      "No PAM50 adjustment: all evaluable PDXs are ",
      unique(analysis_table$PAM50_subtype)
    )

    model_note <- paste0(
      "PAM50 adjustment was not estimable because all evaluable PDXs were ",
      unique(analysis_table$PAM50_subtype),
      "."
    )
  }

  expression_subset <- expression_matrix[
    ,
    analysis_table$samples,
    drop = FALSE
  ]

  fitted_model <- limma::lmFit(
    expression_subset,
    design_matrix
  )

  fitted_model <- limma::eBayes(
    fitted_model,
    robust = TRUE
  )

  differential_expression <- limma::topTable(
    fitted_model,
    coef = "response_groupSensitive",
    number = Inf,
    sort.by = "none"
  )

  ranking_statistic <- differential_expression$t
  names(ranking_statistic) <- rownames(differential_expression)

  ranking_statistic <- ranking_statistic[
    is.finite(ranking_statistic) &
      !is.na(names(ranking_statistic)) &
      names(ranking_statistic) != ""
  ]

  ranking_statistic <- ranking_statistic +
    rank(names(ranking_statistic), ties.method = "first") * 1e-12

  ranking_statistic <- sort(
    ranking_statistic,
    decreasing = TRUE
  )

  gsea_result <- fgsea::fgseaMultilevel(
    pathways = hallmark_pathways,
    stats = ranking_statistic,
    minSize = 10,
    maxSize = 500,
    eps = 0,
    nproc = 1
  )

  gsea_result <- as.data.frame(gsea_result)

  leading_edge_genes <- vapply(
    gsea_result$leadingEdge,
    function(gene_vector) paste(gene_vector, collapse = ";"),
    character(1)
  )

  results <- data.frame(
    treatment = treatment_name,
    response_comparison = "Sensitive versus Resistant",
    pathway_gene_set = gsea_result$pathway,
    pathway_label = gsub("^HALLMARK_", "", gsea_result$pathway),
    NES = gsea_result$NES,
    p_value = gsea_result$pval,
    FDR_q_within_treatment = p.adjust(
      gsea_result$pval,
      method = "BH"
    ),
    FDR_q_global = NA_real_,
    n_resistant = n_resistant,
    n_sensitive = n_sensitive,
    n_total = n_total,
    n_PAM50_subtypes = n_pam50_subtypes,
    adjustment = adjustment_label,
    model_status = "success",
    leading_edge_genes = leading_edge_genes,
    stringsAsFactors = FALSE
  )

  model_summary <- data.frame(
    treatment = treatment_name,
    metadata_column = response_column,
    n_total = n_total,
    n_resistant = n_resistant,
    n_sensitive = n_sensitive,
    n_PAM50_subtypes = n_pam50_subtypes,
    PAM50_subtypes_present = paste(
      sort(unique(analysis_table$PAM50_subtype)),
      collapse = "; "
    ),
    adjustment = adjustment_label,
    model_status = "success",
    note = model_note,
    stringsAsFactors = FALSE
  )

  analysis_table$treatment <- treatment_name
  analysis_table$metadata_column <- response_column

  list(
    results = results,
    summary = model_summary,
    input = analysis_table
  )
}
```

```{r run-all-treatment-cohorts}
analysis_list <- lapply(
  seq_len(nrow(treatment_definitions)),
  function(treatment_index) {
    run_treatment_gsea(
      treatment_name = treatment_definitions$treatment[treatment_index],
      response_column = treatment_definitions$metadata_column[treatment_index],
      metadata = metadata,
      pam50_calls = pam50_calls,
      expression_matrix = expression_log2cpm,
      hallmark_pathways = hallmark_pathways
    )
  }
)

all_gsea_results <- do.call(
  rbind,
  lapply(analysis_list, function(current_analysis) {
    current_analysis$results
  })
)

model_summary <- do.call(
  rbind,
  lapply(analysis_list, function(current_analysis) {
    current_analysis$summary
  })
)

input_annotations <- do.call(
  rbind,
  lapply(analysis_list, function(current_analysis) {
    current_analysis$input
  })
)

all_gsea_results$FDR_q_global <- p.adjust(
  all_gsea_results$p_value,
  method = "BH"
)

all_gsea_results <- all_gsea_results[
  order(
    all_gsea_results$FDR_q_global,
    all_gsea_results$p_value,
    -abs(all_gsea_results$NES)
  ),
  ,
  drop = FALSE
]

significant_gsea_results <- all_gsea_results[
  all_gsea_results$FDR_q_global < 0.05,
  ,
  drop = FALSE
]
```

# Export results

```{r export-results}
write.csv(
  all_gsea_results,
  file.path(results_dir, "Hallmark_GSEA_all_results.csv"),
  row.names = FALSE,
  quote = TRUE
)

write.csv(
  model_summary,
  file.path(results_dir, "Hallmark_GSEA_model_summary.csv"),
  row.names = FALSE,
  quote = TRUE
)

write.csv(
  input_annotations,
  file.path(results_dir, "Hallmark_GSEA_input_annotations.csv"),
  row.names = FALSE,
  quote = TRUE
)

write.csv(
  significant_gsea_results,
  file.path(results_dir, "Hallmark_GSEA_global_FDR_significant.csv"),
  row.names = FALSE,
  quote = TRUE
)

model_summary
```

# Reproducibility information

```{r session-information}
sessionInfo()
```
