---
title: "Continuous and joint-response GSEA for olaparib and cisplatin"
author: "Andreu Odena"
date: "2026-09-08"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Objective

This analysis evaluates baseline transcriptional programmes associated with
continuous olaparib and cisplatin benefit and compares the joint response
groups defined from their parallel *in vivo* evaluations. All gene-level
models are adjusted for Parker PAM50 intrinsic subtype.

For the continuous analyses, tumour-volume change is multiplied by -1 so that
higher values indicate greater therapeutic benefit. Positive enrichment scores
therefore indicate higher baseline expression with greater benefit. For the
joint-response comparisons, positive enrichment scores indicate higher
expression in the first group named in the comparison.

No figures are generated by this R Markdown.

# Packages and paths

The analysis should be run from the root of the GitHub repository. PAM50 calls
are generated by `02_PAM50_classification.Rmd`.

```{r packages-and-paths}
library(edgeR)
library(limma)
library(fgsea)
library(msigdbr)

counts_path <- file.path("data", "260824_Counts.csv")
response_metadata_path <- file.path(
  "data",
  "260827_Metadata_PARPi_Cisplatin.csv"
)
pam50_path <- file.path(
  "results",
  "PAM50",
  "PAM50_Parker_calls.csv"
)

results_dir <- file.path(
  "results",
  "olaparib_cisplatin_GSEA"
)

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

set.seed(260827)
```

# Load the input data

```{r load-input-data}
counts_input <- read.csv(
  counts_path,
  sep = ";",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

response_metadata <- read.csv(
  response_metadata_path,
  sep = ";",
  check.names = FALSE,
  stringsAsFactors = FALSE,
  fileEncoding = "UTF-8-BOM"
)

pam50_calls <- read.csv(
  pam50_path,
  check.names = FALSE,
  stringsAsFactors = FALSE,
  fileEncoding = "UTF-8-BOM"
)

gene_names <- toupper(trimws(as.character(counts_input$gene_name)))
counts_matrix <- as.matrix(counts_input[, -1, drop = FALSE])
storage.mode(counts_matrix) <- "numeric"

valid_gene <- !is.na(gene_names) & gene_names != ""
counts_matrix <- counts_matrix[valid_gene, , drop = FALSE]
gene_names <- gene_names[valid_gene]

counts_matrix <- rowsum(
  counts_matrix,
  group = gene_names,
  reorder = FALSE
)

response_metadata$samples <- trimws(
  as.character(response_metadata$samples)
)
pam50_calls$samples <- trimws(as.character(pam50_calls$samples))

response_metadata <- merge(
  response_metadata,
  pam50_calls[, c("samples", "PAM50_label")],
  by = "samples",
  all.x = TRUE,
  sort = FALSE
)

colnames(response_metadata)[
  colnames(response_metadata) == "PAM50_label"
] <- "PAM50_subtype"

response_metadata$best_response_olap <- as.numeric(
  gsub(
    ",",
    ".",
    trimws(as.character(response_metadata$best_response_olap)),
    fixed = TRUE
  )
)

response_metadata$best_response_cisplatin <- as.numeric(
  gsub(
    ",",
    ".",
    trimws(as.character(response_metadata$best_response_cisplatin)),
    fixed = TRUE
  )
)
```

# MSigDB gene sets

Hallmark and Reactome gene sets are retrieved from MSigDB. The separate
continuous drug analyses use both collections, whereas the joint-response
analyses use Reactome pathways.

```{r msigdb-gene-sets}
hallmark_table <- msigdbr::msigdbr(
  species = "Homo sapiens",
  collection = "H"
)

reactome_table <- msigdbr::msigdbr(
  species = "Homo sapiens",
  collection = "C2",
  subcollection = "CP:REACTOME"
)

hallmark_pathways <- split(
  toupper(hallmark_table$gene_symbol),
  hallmark_table$gs_name
)

reactome_pathways <- split(
  toupper(reactome_table$gene_symbol),
  reactome_table$gs_name
)

hallmark_pathways <- lapply(hallmark_pathways, unique)
reactome_pathways <- lapply(reactome_pathways, unique)
```

# Analysis functions

```{r analysis-functions}
fit_gene_ranking <- function(
    sample_metadata,
    association_variable,
    counts_matrix,
    filtering = "fraction") {

  sample_metadata$PAM50_subtype <- droplevels(
    factor(sample_metadata$PAM50_subtype)
  )
  sample_metadata$association_variable <- association_variable

  analysis_counts <- counts_matrix[
    ,
    sample_metadata$samples,
    drop = FALSE
  ]

  design <- model.matrix(
    ~ association_variable + PAM50_subtype,
    data = sample_metadata
  )

  dge <- DGEList(counts = analysis_counts)

  if (filtering == "fraction") {
    minimum_samples <- ceiling(0.20 * ncol(analysis_counts))
    keep_gene <- rowSums(analysis_counts > 10) >= minimum_samples
  } else {
    keep_gene <- filterByExpr(dge, design = design)
  }

  dge <- dge[keep_gene, , keep.lib.sizes = FALSE]
  dge <- calcNormFactors(dge, method = "TMM")

  voom_object <- voom(dge, design = design, plot = FALSE)
  fit <- lmFit(voom_object, design)
  fit <- eBayes(fit, trend = TRUE, robust = TRUE)

  gene_ranking <- topTable(
    fit,
    coef = "association_variable",
    number = Inf,
    sort.by = "none"
  )

  gene_ranking$gene_symbol <- rownames(gene_ranking)
  gene_ranking <- gene_ranking[
    order(gene_ranking$t, decreasing = TRUE),
    ,
    drop = FALSE
  ]

  ranked_statistics <- gene_ranking$t
  names(ranked_statistics) <- gene_ranking$gene_symbol
  ranked_statistics <- ranked_statistics[
    is.finite(ranked_statistics) &
      !is.na(names(ranked_statistics)) &
      names(ranked_statistics) != ""
  ]

  ranked_statistics <- ranked_statistics +
    rank(names(ranked_statistics), ties.method = "first") * 1e-12
  ranked_statistics <- sort(ranked_statistics, decreasing = TRUE)

  list(
    gene_ranking = gene_ranking,
    ranked_statistics = ranked_statistics,
    n_models = nrow(sample_metadata),
    n_genes = nrow(gene_ranking)
  )
}

run_preranked_gsea <- function(
    ranked_statistics,
    pathways,
    collection,
    analysis_id,
    comparison,
    positive_direction,
    negative_direction) {

  set.seed(260827)

  gsea_results <- fgseaMultilevel(
    pathways = pathways,
    stats = ranked_statistics,
    minSize = 15,
    maxSize = 500,
    eps = 0,
    nproc = 1
  )

  gsea_results <- as.data.frame(gsea_results)
  gsea_results$leadingEdge <- vapply(
    gsea_results$leadingEdge,
    paste,
    collapse = ";",
    FUN.VALUE = character(1)
  )

  gsea_results$analysis_id <- analysis_id
  gsea_results$comparison <- comparison
  gsea_results$collection <- collection
  gsea_results$direction <- ifelse(
    gsea_results$NES > 0,
    positive_direction,
    negative_direction
  )
  gsea_results$FDR_q_within_collection <- p.adjust(
    gsea_results$pval,
    method = "BH"
  )

  gsea_results
}

make_pathway_label <- function(pathway_name) {
  pathway_label <- gsub("^HALLMARK_|^REACTOME_", "", pathway_name)
  pathway_label <- gsub("_", " ", pathway_label)
  tools::toTitleCase(tolower(pathway_label))
}
```

# Continuous response to each drug

For each drug, the best tumour-volume response is converted to a standardized
benefit score. Gene expression is modelled as a function of this score and
PAM50 subtype. Limma moderated *t*-statistics are used to rank genes for
Hallmark and Reactome GSEA.

```{r continuous-single-drug-gsea}
continuous_definitions <- list(
  list(
    analysis_id = "continuous_olaparib",
    drug = "Olaparib",
    response_column = "best_response_olap"
  ),
  list(
    analysis_id = "continuous_cisplatin",
    drug = "Cisplatin",
    response_column = "best_response_cisplatin"
  )
)

continuous_results <- list()
continuous_rankings <- list()
continuous_annotations <- list()
continuous_summaries <- list()

for (definition in continuous_definitions) {

  analysis_metadata <- response_metadata[
    !is.na(response_metadata[[definition$response_column]]) &
      !is.na(response_metadata$PAM50_subtype) &
      response_metadata$PAM50_subtype != "" &
      response_metadata$samples %in% colnames(counts_matrix),
    ,
    drop = FALSE
  ]

  analysis_metadata <- analysis_metadata[
    match(
      intersect(colnames(counts_matrix), analysis_metadata$samples),
      analysis_metadata$samples
    ),
    ,
    drop = FALSE
  ]

  best_response <- analysis_metadata[[definition$response_column]]
  benefit <- -best_response
  benefit_z <- as.numeric(scale(benefit))

  model_result <- fit_gene_ranking(
    sample_metadata = analysis_metadata,
    association_variable = benefit_z,
    counts_matrix = counts_matrix,
    filtering = "fraction"
  )

  hallmark_result <- run_preranked_gsea(
    ranked_statistics = model_result$ranked_statistics,
    pathways = hallmark_pathways,
    collection = "Hallmark",
    analysis_id = definition$analysis_id,
    comparison = paste0("Continuous ", definition$drug, " benefit"),
    positive_direction = paste0(
      "Higher expression with greater ",
      tolower(definition$drug),
      " benefit"
    ),
    negative_direction = paste0(
      "Higher expression with lower ",
      tolower(definition$drug),
      " benefit"
    )
  )

  reactome_result <- run_preranked_gsea(
    ranked_statistics = model_result$ranked_statistics,
    pathways = reactome_pathways,
    collection = "Reactome",
    analysis_id = definition$analysis_id,
    comparison = paste0("Continuous ", definition$drug, " benefit"),
    positive_direction = paste0(
      "Higher expression with greater ",
      tolower(definition$drug),
      " benefit"
    ),
    negative_direction = paste0(
      "Higher expression with lower ",
      tolower(definition$drug),
      " benefit"
    )
  )

  drug_results <- rbind(hallmark_result, reactome_result)
  drug_results$FDR_q_global_Hallmark_Reactome <- p.adjust(
    drug_results$pval,
    method = "BH"
  )
  drug_results$pathway_label <- make_pathway_label(
    drug_results$pathway
  )

  model_result$gene_ranking$analysis_id <- definition$analysis_id
  model_result$gene_ranking$direction <- ifelse(
    model_result$gene_ranking$t > 0,
    paste0(
      "Higher expression with greater ",
      tolower(definition$drug),
      " benefit"
    ),
    paste0(
      "Higher expression with lower ",
      tolower(definition$drug),
      " benefit"
    )
  )

  continuous_results[[definition$analysis_id]] <- drug_results
  continuous_rankings[[definition$analysis_id]] <-
    model_result$gene_ranking

  continuous_annotations[[definition$analysis_id]] <- data.frame(
    analysis_id = definition$analysis_id,
    samples = analysis_metadata$samples,
    PAM50_subtype = analysis_metadata$PAM50_subtype,
    best_response_percent_change = best_response,
    benefit = benefit,
    benefit_z = benefit_z,
    stringsAsFactors = FALSE
  )

  continuous_summaries[[definition$analysis_id]] <- data.frame(
    analysis_id = definition$analysis_id,
    drug = definition$drug,
    n_models = model_result$n_models,
    n_genes_ranked = model_result$n_genes,
    model = "Expression ~ standardized benefit + PAM50 subtype",
    positive_NES = paste0(
      "Higher baseline expression with greater ",
      tolower(definition$drug),
      " benefit"
    ),
    stringsAsFactors = FALSE
  )
}

continuous_results <- do.call(rbind, continuous_results)
continuous_rankings <- do.call(rbind, continuous_rankings)
continuous_annotations <- do.call(rbind, continuous_annotations)
continuous_summaries <- do.call(rbind, continuous_summaries)

continuous_results <- continuous_results[
  order(
    continuous_results$analysis_id,
    continuous_results$FDR_q_global_Hallmark_Reactome,
    continuous_results$pval
  ),
  ,
  drop = FALSE
]

write.csv(
  continuous_results,
  file.path(results_dir, "Continuous_GSEA_all_results.csv"),
  row.names = FALSE
)

write.csv(
  continuous_results[
    continuous_results$FDR_q_global_Hallmark_Reactome < 0.05,
    ,
    drop = FALSE
  ],
  file.path(results_dir, "Continuous_GSEA_global_FDR_significant.csv"),
  row.names = FALSE
)

write.csv(
  continuous_rankings,
  file.path(results_dir, "Continuous_GSEA_gene_rankings.csv"),
  row.names = FALSE
)

write.csv(
  continuous_annotations,
  file.path(results_dir, "Continuous_GSEA_input_annotations.csv"),
  row.names = FALSE
)

write.csv(
  continuous_summaries,
  file.path(results_dir, "Continuous_GSEA_model_summary.csv"),
  row.names = FALSE
)
```

# Joint olaparib-cisplatin response groups

Progressive disease is classified as resistance. Stable disease, partial
response and complete response are classified as clinical benefit. When the
categorical annotation is missing, the quantitative best response is used and
values above 20% are classified as progressive disease.

The four joint groups are dual response, olaparib-selective response,
cisplatin-selective response and dual resistance. The olaparib-selective group
is retained in the exported annotation but is not included in GSEA because it
contains only two models.

```{r joint-response-definitions}
make_binary_response <- function(response_values, best_response) {
  cleaned_values <- toupper(trimws(as.character(response_values)))
  cleaned_values[cleaned_values %in% c(
    "", "NA", "N/A", "NOT TESTED", "NONE"
  )] <- NA_character_

  binary_response <- rep(NA_character_, length(cleaned_values))
  binary_response[cleaned_values %in% "PD"] <- "Resistant"
  binary_response[cleaned_values %in% c(
    "SD", "PR", "CR", "SD/PR/CR"
  )] <- "Sensitive"

  use_best_response <- is.na(binary_response) & !is.na(best_response)
  binary_response[
    use_best_response & best_response > 20
  ] <- "Resistant"
  binary_response[
    use_best_response & best_response <= 20
  ] <- "Sensitive"

  binary_response
}

joint_metadata <- response_metadata

joint_metadata$Olaparib_response <- make_binary_response(
  joint_metadata$`Olaparib_(PARP1-2i)_response`,
  joint_metadata$best_response_olap
)

joint_metadata$Cisplatin_response <- make_binary_response(
  joint_metadata$`Cisplatin_(platin_chemo)_response`,
  joint_metadata$best_response_cisplatin
)

joint_metadata$Joint_response_group <- NA_character_

joint_metadata$Joint_response_group[
  joint_metadata$Olaparib_response == "Sensitive" &
    joint_metadata$Cisplatin_response == "Sensitive"
] <- "Dual_response"

joint_metadata$Joint_response_group[
  joint_metadata$Olaparib_response == "Sensitive" &
    joint_metadata$Cisplatin_response == "Resistant"
] <- "Olaparib_selective_response"

joint_metadata$Joint_response_group[
  joint_metadata$Olaparib_response == "Resistant" &
    joint_metadata$Cisplatin_response == "Sensitive"
] <- "Cisplatin_selective_response"

joint_metadata$Joint_response_group[
  joint_metadata$Olaparib_response == "Resistant" &
    joint_metadata$Cisplatin_response == "Resistant"
] <- "Dual_resistance"

joint_metadata <- joint_metadata[
  !is.na(joint_metadata$Joint_response_group) &
    !is.na(joint_metadata$PAM50_subtype) &
    joint_metadata$PAM50_subtype != "" &
    joint_metadata$samples %in% colnames(counts_matrix),
  ,
  drop = FALSE
]

joint_metadata <- joint_metadata[
  match(
    intersect(colnames(counts_matrix), joint_metadata$samples),
    joint_metadata$samples
  ),
  ,
  drop = FALSE
]

write.csv(
  joint_metadata[, c(
    "samples",
    "PAM50_subtype",
    "best_response_olap",
    "best_response_cisplatin",
    "Olaparib_response",
    "Cisplatin_response",
    "Joint_response_group"
  )],
  file.path(results_dir, "Joint_response_input_annotations.csv"),
  row.names = FALSE
)

joint_group_counts <- as.data.frame(
  table(joint_metadata$Joint_response_group),
  stringsAsFactors = FALSE
)
colnames(joint_group_counts) <- c("Joint_response_group", "n_models")

write.csv(
  joint_group_counts,
  file.path(results_dir, "Joint_response_group_counts.csv"),
  row.names = FALSE
)
```

# Reactome GSEA between joint-response groups

Three pairwise comparisons are performed. Positive enrichment scores indicate
higher expression in the first group named in each comparison. Within-
comparison FDR and FDR across all pathways from all three comparisons are
reported.

```{r joint-response-reactome-gsea}
group_labels <- c(
  Dual_response = "dual response",
  Cisplatin_selective_response = "cisplatin-selective response",
  Dual_resistance = "dual resistance"
)

comparison_definitions <- list(
  list(
    comparison_id = "dual_response_vs_dual_resistance",
    comparison = "Dual response versus dual resistance",
    positive_group = "Dual_response",
    negative_group = "Dual_resistance"
  ),
  list(
    comparison_id = "cisplatin_selective_vs_dual_resistance",
    comparison =
      "Cisplatin-selective response versus dual resistance",
    positive_group = "Cisplatin_selective_response",
    negative_group = "Dual_resistance"
  ),
  list(
    comparison_id = "cisplatin_selective_vs_dual_response",
    comparison =
      "Cisplatin-selective response versus dual response",
    positive_group = "Cisplatin_selective_response",
    negative_group = "Dual_response"
  )
)

joint_results <- list()
joint_rankings <- list()
joint_summaries <- list()

for (definition in comparison_definitions) {

  comparison_metadata <- joint_metadata[
    joint_metadata$Joint_response_group %in% c(
      definition$positive_group,
      definition$negative_group
    ),
    ,
    drop = FALSE
  ]

  response_indicator <- as.integer(
    comparison_metadata$Joint_response_group ==
      definition$positive_group
  )

  model_result <- fit_gene_ranking(
    sample_metadata = comparison_metadata,
    association_variable = response_indicator,
    counts_matrix = counts_matrix,
    filtering = "filterByExpr"
  )

  positive_label <- group_labels[[definition$positive_group]]
  negative_label <- group_labels[[definition$negative_group]]

  comparison_result <- run_preranked_gsea(
    ranked_statistics = model_result$ranked_statistics,
    pathways = reactome_pathways,
    collection = "Reactome",
    analysis_id = definition$comparison_id,
    comparison = definition$comparison,
    positive_direction = paste0(
      "Higher expression with ",
      positive_label
    ),
    negative_direction = paste0(
      "Higher expression with ",
      negative_label
    )
  )

  comparison_result$comparison_id <- definition$comparison_id
  comparison_result$pathway_label <- make_pathway_label(
    comparison_result$pathway
  )
  comparison_result$FDR_q_within_comparison <-
    comparison_result$FDR_q_within_collection
  comparison_result$n_positive <- sum(response_indicator == 1)
  comparison_result$n_negative <- sum(response_indicator == 0)

  model_result$gene_ranking$comparison_id <-
    definition$comparison_id
  model_result$gene_ranking$comparison <- definition$comparison
  model_result$gene_ranking$direction <- ifelse(
    model_result$gene_ranking$t > 0,
    paste0("Higher expression with ", positive_label),
    paste0("Higher expression with ", negative_label)
  )

  joint_results[[definition$comparison_id]] <- comparison_result
  joint_rankings[[definition$comparison_id]] <-
    model_result$gene_ranking

  joint_summaries[[definition$comparison_id]] <- data.frame(
    comparison_id = definition$comparison_id,
    comparison = definition$comparison,
    positive_group = definition$positive_group,
    negative_group = definition$negative_group,
    n_positive = sum(response_indicator == 1),
    n_negative = sum(response_indicator == 0),
    n_genes_ranked = model_result$n_genes,
    model = "Expression ~ response group + PAM50 subtype",
    stringsAsFactors = FALSE
  )
}

joint_results <- as.data.frame(
  do.call(rbind, lapply(joint_results, as.data.frame)),
  stringsAsFactors = FALSE
)
joint_rankings <- as.data.frame(
  do.call(rbind, lapply(joint_rankings, as.data.frame)),
  stringsAsFactors = FALSE
)
joint_summaries <- as.data.frame(
  do.call(rbind, lapply(joint_summaries, as.data.frame)),
  stringsAsFactors = FALSE
)

joint_results[["comparison_id"]] <- as.character(
  joint_results[["comparison_id"]]
)
joint_results[["FDR_q_within_comparison"]] <- as.numeric(
  joint_results[["FDR_q_within_comparison"]]
)
joint_results[["NES"]] <- as.numeric(joint_results[["NES"]])
joint_results[["pval"]] <- as.numeric(joint_results[["pval"]])

joint_results$FDR_q_across_all_three_comparisons <- p.adjust(
  joint_results[["pval"]],
  method = "BH"
)

joint_order <- order(
  joint_results[["comparison_id"]],
  joint_results[["FDR_q_within_comparison"]],
  -abs(joint_results[["NES"]])
)

joint_results <- joint_results[
  joint_order,
  ,
  drop = FALSE
]

write.csv(
  joint_results,
  file.path(results_dir, "Joint_response_Reactome_GSEA_all_results.csv"),
  row.names = FALSE
)

write.csv(
  joint_results[
    joint_results$FDR_q_within_comparison <= 0.10,
    ,
    drop = FALSE
  ],
  file.path(
    results_dir,
    "Joint_response_Reactome_GSEA_within_comparison_FDR_q10.csv"
  ),
  row.names = FALSE
)

write.csv(
  joint_results[
    joint_results$FDR_q_across_all_three_comparisons <= 0.10,
    ,
    drop = FALSE
  ],
  file.path(
    results_dir,
    "Joint_response_Reactome_GSEA_global_FDR_q10.csv"
  ),
  row.names = FALSE
)

write.csv(
  joint_rankings,
  file.path(results_dir, "Joint_response_gene_rankings.csv"),
  row.names = FALSE
)

write.csv(
  joint_summaries,
  file.path(results_dir, "Joint_response_model_summary.csv"),
  row.names = FALSE
)
```

# Session information

```{r session-information}
sessionInfo()
```
