---
title: "Paired transcriptional response to olaparib in sensitive and acquired-resistant PDXs"
author: "Andreu Odena"
date: "2026-09-08"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Objective

This analysis identifies transcriptional programmes regulated by olaparib in
seven baseline PARPi-sensitive PDX models and their seven acquired-resistant
counterparts. Olaparib-treated and untreated samples are compared using paired
models within each resistance state. Limma moderated *t*-statistics are used to
rank genes for Hallmark and Reactome GSEA.

Positive enrichment scores indicate higher expression after olaparib treatment.
Negative enrichment scores indicate higher expression in untreated samples.
No figures are generated by this R Markdown.

# Packages and paths

```{r packages-and-paths}
library(edgeR)
library(limma)
library(fgsea)
library(msigdbr)

counts_path <- file.path(
  "data",
  "COUNTS_PARPi_Resistance_Cohort.csv"
)

metadata_path <- file.path(
  "data",
  "Metadata_PARPi_Resistance_Cohort.csv"
)

results_dir <- file.path(
  "results",
  "paired_PARPi_acquired_resistance_GSEA"
)

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

set.seed(260901)
```

# Load sample annotations

The analysis contains two resistance states and two treatment conditions. The
`pdx_id` variable identifies the paired untreated and olaparib-treated samples
within each state.

```{r load-sample-annotations}
metadata <- read.csv(
  metadata_path,
  sep = ";",
  check.names = FALSE,
  stringsAsFactors = FALSE,
  fileEncoding = "UTF-8-BOM"
)

metadata$sample_id <- trimws(as.character(metadata$sample_id))
metadata$pdx_id <- trimws(as.character(metadata$pdx_id))
metadata$patient_id <- trimws(as.character(metadata$patient_id))

metadata$resistance_state <- ifelse(
  tolower(trimws(metadata$olaparib_sensitivity)) == "sensitive",
  "Baseline_PARPi_sensitive",
  "Acquired_PARPi_resistant"
)

metadata$treatment_group <- ifelse(
  tolower(trimws(metadata$treatment_status)) == "olaparib",
  "Olaparib_treated",
  "Untreated"
)

metadata$treatment_group <- factor(
  metadata$treatment_group,
  levels = c("Untreated", "Olaparib_treated")
)
```

# Load human RNA-seq counts

Only human genes, identified by Ensembl gene identifiers beginning with
`ENSG`, are retained. Raw counts assigned to duplicate gene symbols are summed.

```{r load-human-counts}
counts_input <- read.csv(
  counts_path,
  sep = ";",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

human_rows <- grepl("^ENSG", counts_input$id_gene)
counts_human <- counts_input[human_rows, , drop = FALSE]

gene_symbols <- toupper(
  trimws(as.character(counts_human$gene_name))
)

valid_gene <- !is.na(gene_symbols) & gene_symbols != ""
counts_human <- counts_human[valid_gene, , drop = FALSE]
gene_symbols <- gene_symbols[valid_gene]

sample_columns <- metadata$sample_id

counts_matrix <- as.matrix(
  counts_human[, sample_columns, drop = FALSE]
)
storage.mode(counts_matrix) <- "numeric"

counts_matrix <- rowsum(
  counts_matrix,
  group = gene_symbols,
  reorder = FALSE
)
```

# Hallmark and Reactome gene sets

```{r msigdb-gene-sets}
hallmark_table <- msigdbr::msigdbr(
  species = "Homo sapiens",
  collection = "H"
)

reactome_table <- msigdbr::msigdbr(
  species = "Homo sapiens",
  collection = "C2",
  subcollection = "CP:REACTOME"
)

hallmark_pathways <- split(
  toupper(hallmark_table$gene_symbol),
  hallmark_table$gs_name
)

reactome_pathways <- split(
  toupper(reactome_table$gene_symbol),
  reactome_table$gs_name
)

hallmark_pathways <- lapply(hallmark_pathways, unique)
reactome_pathways <- lapply(reactome_pathways, unique)
```

# Analysis functions

```{r analysis-functions}
make_pathway_label <- function(pathway_name) {
  pathway_label <- gsub("^HALLMARK_|^REACTOME_", "", pathway_name)
  pathway_label <- gsub("_", " ", pathway_label)
  tools::toTitleCase(tolower(pathway_label))
}

run_paired_expression_model <- function(
    resistance_state,
    contrast_id,
    metadata,
    counts_matrix) {

  analysis_metadata <- metadata[
    metadata$resistance_state == resistance_state,
    ,
    drop = FALSE
  ]

  pair_table <- table(
    analysis_metadata$pdx_id,
    analysis_metadata$treatment_group
  )

  complete_pairs <- rownames(pair_table)[
    apply(pair_table > 0, 1, all)
  ]

  analysis_metadata <- analysis_metadata[
    analysis_metadata$pdx_id %in% complete_pairs,
    ,
    drop = FALSE
  ]

  analysis_metadata <- analysis_metadata[
    order(
      analysis_metadata$pdx_id,
      analysis_metadata$treatment_group
    ),
    ,
    drop = FALSE
  ]

  analysis_counts <- counts_matrix[
    ,
    analysis_metadata$sample_id,
    drop = FALSE
  ]

  analysis_metadata$pdx_id <- factor(analysis_metadata$pdx_id)

  design <- model.matrix(
    ~ 0 + pdx_id + treatment_group,
    data = analysis_metadata
  )

  dge <- DGEList(counts = analysis_counts)
  keep_gene <- filterByExpr(dge, design = design)
  dge <- dge[keep_gene, , keep.lib.sizes = FALSE]
  dge <- calcNormFactors(dge, method = "TMM")

  voom_object <- voom(dge, design = design, plot = FALSE)
  fit <- lmFit(voom_object, design)
  fit <- eBayes(fit, robust = TRUE)

  gene_results <- topTable(
    fit,
    coef = "treatment_groupOlaparib_treated",
    number = Inf,
    sort.by = "none"
  )

  gene_results$gene_symbol <- rownames(gene_results)
  gene_results$contrast_id <- contrast_id
  gene_results$direction <- ifelse(
    gene_results$logFC > 0,
    "Higher expression after olaparib treatment",
    "Higher expression in untreated samples"
  )

  gene_results <- gene_results[
    order(gene_results$t, decreasing = TRUE),
    ,
    drop = FALSE
  ]

  ranked_statistics <- gene_results$t
  names(ranked_statistics) <- gene_results$gene_symbol
  ranked_statistics <- ranked_statistics[
    is.finite(ranked_statistics) &
      !is.na(names(ranked_statistics)) &
      names(ranked_statistics) != ""
  ]

  ranked_statistics <- ranked_statistics +
    rank(names(ranked_statistics), ties.method = "first") * 1e-12
  ranked_statistics <- sort(ranked_statistics, decreasing = TRUE)

  list(
    metadata = analysis_metadata,
    gene_results = gene_results,
    ranked_statistics = ranked_statistics,
    n_pairs = length(unique(analysis_metadata$pdx_id)),
    n_samples = nrow(analysis_metadata),
    n_genes = nrow(gene_results)
  )
}

run_paired_gsea <- function(
    ranked_statistics,
    pathways,
    collection,
    contrast_id,
    resistance_state) {

  set.seed(260901)

  gsea_results <- fgseaMultilevel(
    pathways = pathways,
    stats = ranked_statistics,
    minSize = 15,
    maxSize = 500,
    eps = 0,
    nproc = 1
  )

  gsea_results <- as.data.frame(gsea_results)
  gsea_results$leadingEdge <- vapply(
    gsea_results$leadingEdge,
    paste,
    collapse = ";",
    FUN.VALUE = character(1)
  )

  gsea_results$contrast_id <- contrast_id
  gsea_results$resistance_state <- resistance_state
  gsea_results$collection <- collection
  gsea_results$pathway_label <- make_pathway_label(
    gsea_results$pathway
  )
  gsea_results$direction <- ifelse(
    gsea_results$NES > 0,
    "Higher expression after olaparib treatment",
    "Higher expression in untreated samples"
  )
  gsea_results$FDR_q_within_collection <- p.adjust(
    gsea_results$pval,
    method = "BH"
  )

  gsea_results
}
```

# Paired treatment analyses

The olaparib treatment effect is estimated separately in sensitive and
acquired-resistant PDXs while accounting for each PDX pair in the design
matrix.

```{r paired-treatment-models}
analysis_definitions <- list(
  list(
    resistance_state = "Baseline_PARPi_sensitive",
    contrast_id =
      "Olaparib_treated_vs_untreated_in_PARPi_sensitive"
  ),
  list(
    resistance_state = "Acquired_PARPi_resistant",
    contrast_id =
      "Olaparib_treated_vs_untreated_in_acquired_resistance"
  )
)

expression_results <- list()
gsea_results <- list()
analysis_annotations <- list()
model_summaries <- list()

for (definition in analysis_definitions) {

  paired_model <- run_paired_expression_model(
    resistance_state = definition$resistance_state,
    contrast_id = definition$contrast_id,
    metadata = metadata,
    counts_matrix = counts_matrix
  )

  hallmark_result <- run_paired_gsea(
    ranked_statistics = paired_model$ranked_statistics,
    pathways = hallmark_pathways,
    collection = "Hallmark",
    contrast_id = definition$contrast_id,
    resistance_state = definition$resistance_state
  )

  reactome_result <- run_paired_gsea(
    ranked_statistics = paired_model$ranked_statistics,
    pathways = reactome_pathways,
    collection = "Reactome",
    contrast_id = definition$contrast_id,
    resistance_state = definition$resistance_state
  )

  expression_results[[definition$contrast_id]] <-
    paired_model$gene_results

  gsea_results[[definition$contrast_id]] <- rbind(
    hallmark_result,
    reactome_result
  )

  analysis_annotations[[definition$contrast_id]] <-
    paired_model$metadata

  model_summaries[[definition$contrast_id]] <- data.frame(
    contrast_id = definition$contrast_id,
    resistance_state = definition$resistance_state,
    n_paired_PDX_models = paired_model$n_pairs,
    n_samples = paired_model$n_samples,
    n_human_genes_tested = paired_model$n_genes,
    model = "Expression ~ PDX pair + olaparib treatment",
    positive_NES = "Higher expression after olaparib treatment",
    stringsAsFactors = FALSE
  )
}
```

# Combine and export results

```{r combine-and-export-results}
all_expression_results <- as.data.frame(
  do.call(rbind, expression_results),
  stringsAsFactors = FALSE
)

all_gsea_results <- as.data.frame(
  do.call(rbind, gsea_results),
  stringsAsFactors = FALSE
)

all_annotations <- as.data.frame(
  do.call(rbind, analysis_annotations),
  stringsAsFactors = FALSE
)

model_summary <- as.data.frame(
  do.call(rbind, model_summaries),
  stringsAsFactors = FALSE
)

all_gsea_results$FDR_q_global_within_contrast <- ave(
  all_gsea_results$pval,
  all_gsea_results$contrast_id,
  FUN = function(p_values) {
    p.adjust(p_values, method = "BH")
  }
)

all_gsea_results$FDR_q_across_all_tests <- p.adjust(
  all_gsea_results$pval,
  method = "BH"
)

all_gsea_results <- all_gsea_results[
  order(
    all_gsea_results$contrast_id,
    all_gsea_results$collection,
    all_gsea_results$FDR_q_within_collection,
    -abs(all_gsea_results$NES)
  ),
  ,
  drop = FALSE
]

write.csv(
  all_annotations,
  file.path(results_dir, "Paired_treatment_sample_annotations.csv"),
  row.names = FALSE
)

write.csv(
  model_summary,
  file.path(results_dir, "Paired_treatment_model_summary.csv"),
  row.names = FALSE
)

write.csv(
  all_expression_results,
  file.path(results_dir, "Paired_treatment_DE_all_genes.csv"),
  row.names = FALSE
)

write.csv(
  all_gsea_results,
  file.path(results_dir, "Paired_treatment_GSEA_all_results.csv"),
  row.names = FALSE
)

write.csv(
  all_gsea_results[
    all_gsea_results$FDR_q_within_collection < 0.05,
    ,
    drop = FALSE
  ],
  file.path(
    results_dir,
    "Paired_treatment_GSEA_within_collection_FDR_significant.csv"
  ),
  row.names = FALSE
)

write.csv(
  all_gsea_results[
    all_gsea_results$FDR_q_across_all_tests < 0.05,
    ,
    drop = FALSE
  ],
  file.path(
    results_dir,
    "Paired_treatment_GSEA_global_FDR_significant.csv"
  ),
  row.names = FALSE
)
```

# Session information

```{r session-information}
sessionInfo()
```
